---
title: "Group 39: An analysis of data on Attendance in Scotland"
author: "Yuchen Xu, Yu Li, Lingke Huang, Sirui Lu"
date: "26/4/2026"
output:
  bookdown::pdf_document2:
    latex_engine: pdflatex
    number_sections: true
    toc_depth: 3
header-includes:
  - \usepackage{float}
  - \floatplacement{table}{H}
  - \floatplacement{figure}{H}
  - \usepackage{booktabs}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.pos = 'H')
dat <- read.csv("scottishData.csv")
library(ggplot2)
library(dplyr)
library(broom)

# Subsets for Section 1
glasgow   <- subset(dat, Council_area == "Glasgow City")
edinburgh <- subset(dat, Council_area == "City of Edinburgh")
elsewhere <- subset(dat, Council_area != "Glasgow City" &
                         Council_area != "City of Edinburgh")

# Region variable for Section 3
dat$Region <- ifelse(dat$Council_area == "Glasgow City", "Glasgow",
              ifelse(dat$Council_area == "City of Edinburgh", "Edinburgh", "Elsewhere"))
dat$Region <- factor(dat$Region, levels = c("Glasgow", "Edinburgh", "Elsewhere"))

# Unified colour palette 
col_gla   <- "#D6604D"   # terracotta  — Glasgow
col_edi   <- "#4393C3"   # steel blue  — Edinburgh
col_else  <- "#66C2A5"   # mint green  — Elsewhere
col_point <- "#525252"   # slate gray  — scatter points
col_line  <- "#2166AC"   # deep blue   — regression line
col_ref   <- "#878787"   # medium gray — reference line

# Use serif font in all plots to match the LaTeX document font
theme_set(theme_minimal(base_family = "serif"))
```

\newpage
# Summary
In this report, we investigate the distribution of school attendance rates across Scottish data zones and their relationship with socio-economic factors. We specifically compare attendance between Glasgow City and the City of Edinburgh, and explore which factors are most strongly associated with attendance across a full sample of $N = 400$ data zones.

The report employs descriptive statistics, confidence intervals, and graphical summaries, together with inferential methods including an F-test for equality of variances, Welch's two-sample t-test, Pearson correlation analysis, simple linear regression, a one-sample t-test, and a $\chi^2$ test of independence.

The findings indicate that Glasgow City has a statistically significantly lower mean attendance rate than the City of Edinburgh, although no significant difference in variance is detected. Correlation analysis and simple linear regression applied to the full sample identify the income deprivation rate (Income\_rate) as the strongest single linear predictor of attendance, explaining approximately $58\%$ of its variation ($R^2 = 0.58$). A further investigation reveals that the income--attendance relationship operates fundamentally differently in Glasgow: given Glasgow's extreme income deprivation, its attendance is actually higher than the non-Glasgow baseline would predict. This "Glasgow effect" reflects a structurally different relationship rather than a simple additional penalty, and a residual correlation scan identifies Employment\_rate as the variable most strongly associated with this pattern.
\newpage


# Introduction
School attendance is a fundamental indicator of educational engagement. Research consistently shows that pupils with high rates of absenteeism achieve lower academic outcomes, face greater difficulties in securing employment, and are more likely to experience social exclusion in later life. In Scotland, school attendance varies considerably across geographic areas, and this variation is thought to be associated with underlying differences in socio-economic conditions rather than individual choice alone. Understanding where attendance is lowest, and which factors are most strongly associated with poor attendance, is therefore of direct importance to education authorities and government policymakers seeking to allocate resources effectively.

This report addresses three objectives. First, it compares the distribution of school attendance rates between Glasgow City and the City of Edinburgh, Scotland's two largest cities, to determine whether the two cities differ significantly in their average attendance levels and in the variability of attendance across their constituent areas. Second, using the full available dataset, it investigates which socio-economic indicators are most strongly linearly associated with attendance, and quantifies that relationship through a simple linear regression model. Third, combining the findings from the first two sections, it examines whether Glasgow's low attendance is fully explained by income deprivation, or whether a residual "Glasgow effect" exists, and explores which additional factors may drive it.

The data used in this report are drawn from the 2020 Scottish Index of Multiple Deprivation (SIMD), a composite measure developed by the Scottish Government to identify concentrations of deprivation across Scotland and to support evidence-based policy-making. The dataset contains observations from a sample of $n = 400$ data zones — small, spatially defined geographic units each typically containing between 500 and 1,000 residents, of which there are 6,976 in Scotland in total. Of the 400 sampled zones, 100 are located in Glasgow City, 100 in the City of Edinburgh, and the remaining 200 are from elsewhere in Scotland. Each data zone is described by 34 variables spanning domains including population, income, employment, health, education, geographic access to services, crime, and housing. The primary response variable of interest is Attendance, defined as the proportion of school-age pupils attending school within a given data zone over the academic year. As a continuous numerical variable taking values in the interval $[0, 1]$, the attendance data in the sample exhibits considerable regional variability, ranging from areas of severe absenteeism to those with near-full attendance records.

Identifying the specific socio-economic indicators most closely linked to low attendance is essential for enabling policymakers to target resources and interventions at the most deprived communities, ultimately helping to narrow the educational inequality gap. More broadly, this analysis contributes to the evidence base for understanding how multiple dimensions of deprivation interact to shape educational outcomes, and whether area-level interventions beyond income support may be needed to address persistent attendance disparities.
\newpage

# Methods
Firstly, we used the methods of comparative analysis for Glasgow and Edinburgh. To compare the attendance rates between Glasgow City ($n_1 = 100$) and the City of Edinburgh ($n_2 = 100$), exploratory graphical techniques were initially employed. Side-by-side boxplots were constructed to visually assess differences in medians and identify potential outliers, while histograms were generated to examine the underlying shape and skewness of each city's attendance distribution.

Following the visual inspection, sample means ($\bar{x}$) and sample variances ($s^2$) were calculated. A $95\%$ confidence interval for each city's population mean was constructed using the $t$-distribution: 
$$\bar{x} \pm t_{\alpha/2,\; n-1} \cdot \frac{s}{\sqrt{n}}$$
and a $95\%$ confidence interval for each city's population variance was constructed using the $\chi^2$-distribution:
$$\left(\frac{(n-1)s^2}{\chi^2_{1-\alpha/2,\; n-1}},\quad \frac{(n-1)s^2}{\chi^2_{\alpha/2,\; n-1}}\right)$$
To formally test for a difference in the population mean attendance rates, Welch's two-sample $t$-test was conducted. This test was selected because it does not assume equal population variances. The hypotheses tested were $H_0: \mu_{\text{Glasgow}} = \mu_{\text{Edinburgh}}$ against the two-sided alternative $H_1: \mu_{\text{Glasgow}} \neq \mu_{\text{Edinburgh}}$. Additionally, to compare the variability between the two cities, an $F$-test was applied to test the ratio of the population variances, evaluating $H_0: \sigma^2_{\text{Glasgow}} / \sigma^2_{\text{Edinburgh}} = 1$ versus $H_1: \sigma^2_{\text{Glasgow}} / \sigma^2_{\text{Edinburgh}} \neq 1$.


Secondly, we conducted correlation and linear regression analysis. For this second stage, all $N = 400$ sampled data zones were utilised to explore the broader relationships between the response variable (Attendance, $y$) and four potential socio-economic predictors (Attainment, Crime Rate, Comparative Illness Factor, and Income Rate) using a pairwise scatterplot matrix. The strength and direction of the linear associations were quantified through the Pearson correlation coefficient: 
$$r = \frac{\sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum_{i=1}^{n} (x_i - \bar{x})^2 \sum_{i=1}^{n} (y_i - \bar{y})^2}}$$
Upon identifying the variable with the highest absolute value of $r$ as the primary predictor, a simple linear regression (SLR) model was estimated using Ordinary Least Squares (OLS). The model was specified as:
$$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i, \qquad \epsilon_i \overset{\text{iid}}{\sim} N(0, \sigma^2)$$
where $\beta_0$ is the intercept and $\beta_1$ is the slope coefficient. The overall goodness-of-fit was evaluated using the coefficient of determination $R^2$. Finally, a plot of the residuals ($\hat{\epsilon}_i$) against the fitted values ($\hat{Y}_i$) was examined to verify the assumptions of linearity and homoscedasticity.

Thirdly, we conducted a further investigation into the Glasgow effect by extending the comparison to three regions, including Glasgow, Edinburgh, and the rest of Scotland. We employed pairwise $F$-tests and Welch's $t$-tests to establish Glasgow's distinctiveness. Separate simple linear regression models of Attendance on Income\_rate were fitted for each region to compare slopes, $R^2$ values, and $95\%$ confidence intervals. To quantify the Glasgow effect, a baseline regression was trained on non-Glasgow data ($n = 300$) and used to predict Glasgow's expected attendance; the residuals ($e_i = \text{Actual}_i - \text{Predicted}_i$) were tested via a one-sample $t$-test ($H_0: \mu_e = 0$). A Pearson correlation scan of these residuals against all available continuous variables was conducted to identify candidate explanatory factors. The top-ranked variable was further examined through a scatter plot, linear regression of the residuals on that variable, and a $\chi^2$ test of independence on a $3 \times 3$ contingency table formed by categorising both the residuals and the candidate variable into terciles.

\newpage


# Results

## Section 1: Summary Statistics and Hypothesis Tests for Glasgow versus Edinburgh
This section compares the distribution of Attendance between Glasgow City ($n_1 = 100$) and the City of Edinburgh ($n_2 = 100$) using graphical summaries, descriptive statistics, confidence intervals, and formal hypothesis tests.

### Exploratory Visualisation

```{r city-boxplot, fig.height=4, fig.width=6, fig.cap='Side-by-side boxplots of Attendance for Glasgow and Edinburgh'}

dat_two <- subset(dat, Council_area %in% c("Glasgow City", "City of Edinburgh"))

ggplot(dat_two, aes(x = Council_area, y = Attendance, fill = Council_area)) +
  geom_boxplot(outlier.shape = 1, outlier.size = 1.5) +
  scale_fill_manual(values = c("City of Edinburgh" = col_edi, "Glasgow City" = col_gla)) +
  labs(x = "City", y = "Attendance") +
  theme(legend.position = "none")
```

The boxplots (Figure \@ref(fig:city-boxplot)) reveal a clear difference: Edinburgh has a noticeably higher median attendance rate than Glasgow. Although Glasgow's overall spread is smaller, its distribution is concentrated towards the lower end of the scale, with at least one outlier suggesting that certain Glasgow data zones experience particularly poor attendance.

```{r city-histogram, fig.height=4, fig.width=6, fig.cap='Histograms of Attendance for Glasgow and Edinburgh'}

ggplot(dat_two, aes(x = Attendance, fill = Council_area)) +
  geom_histogram(bins = 20, alpha = 0.7, color = "white") +
  scale_fill_manual(values = c("City of Edinburgh" = col_edi, "Glasgow City" = col_gla)) +
  facet_wrap(~ Council_area, ncol = 2) +
  labs(x = "Attendance", y = "Frequency") +
  theme(legend.position = "none")
```

The city-level histograms (Figure \@ref(fig:city-histogram)) reinforce this impression: Edinburgh's distribution is tightly concentrated at higher values, whereas Glasgow's is wider and exhibits a noticeable left skew, suggesting that some data zones have substantially below-average attendance rates.

### Descriptive Statistics

```{r mean-var}

n_gla <- sum(!is.na(glasgow$Attendance))
n_edi <- sum(!is.na(edinburgh$Attendance))
mean_gla <- mean(glasgow$Attendance, na.rm = TRUE)
mean_edi <- mean(edinburgh$Attendance, na.rm = TRUE)
var_gla  <- var(glasgow$Attendance, na.rm = TRUE)
var_edi  <- var(edinburgh$Attendance, na.rm = TRUE)

sd_gla   <- sd(glasgow$Attendance, na.rm = TRUE)
sd_edi   <- sd(edinburgh$Attendance, na.rm = TRUE)

stats_table <- data.frame(
  City     = c("Glasgow", "Edinburgh"),
  N        = c(n_gla, n_edi),
  Mean     = round(c(mean_gla, mean_edi), 5),
  SD       = round(c(sd_gla, sd_edi), 5),
  Variance = round(c(var_gla, var_edi), 7)
)

knitr::kable(stats_table,
             caption = "Sample means and variances of Attendance for Glasgow and Edinburgh.",
             align = "c",
             booktabs = TRUE)
```

Glasgow's mean attendance is $\bar{x}_G = 0.7613$, compared with $\bar{x}_E = 0.7938$ for Edinburgh — a gap of approximately 3.2 percentage points (Table \@ref(tab:mean-var)). Glasgow's sample variance ($s^2_G = 0.00714$) is smaller than Edinburgh's ($s^2_E = 0.00967$), suggesting slightly less overall dispersion, though the histograms in Figure \@ref(fig:city-histogram) show that Glasgow's distribution has a heavier lower tail.


### Confidence Intervals for Means and Variances
```{r ci-mean}

ci_mean_gla <- mean_gla + c(-1, 1) * qt(0.975, df = n_gla - 1) * sqrt(var_gla / n_gla)
ci_mean_edi <- mean_edi + c(-1, 1) * qt(0.975, df = n_edi - 1) * sqrt(var_edi / n_edi)

ci_mean_table <- data.frame(
  City  = c("Glasgow", "Edinburgh"),
  Lower = round(c(ci_mean_gla[1], ci_mean_edi[1]), 5),
  Upper = round(c(ci_mean_gla[2], ci_mean_edi[2]), 5)
)

knitr::kable(ci_mean_table,
             caption = "95\\% confidence intervals for the mean of Attendance.",
             align = "c",
             col.names = c("City", "2.5%", "97.5%"),
             booktabs = TRUE)
```

The $95\%$ confidence intervals for the population mean (Table \@ref(tab:ci-mean)) are $(0.745, 0.778)$ for Glasgow and $(0.774, 0.813)$ for Edinburgh. These two intervals do not overlap, providing preliminary visual evidence that the true mean attendance rates may differ between the two cities; however, a formal hypothesis test is required to confirm this (see Section 4.1.4). 

```{r ci-var, echo = FALSE}

# 95% CI for variance using chi-squared distribution
ci_var_gla <- c((n_gla - 1) * var_gla / qchisq(0.975, df = n_gla - 1),
                (n_gla - 1) * var_gla / qchisq(0.025, df = n_gla - 1))
ci_var_edi <- c((n_edi - 1) * var_edi / qchisq(0.975, df = n_edi - 1),
                (n_edi - 1) * var_edi / qchisq(0.025, df = n_edi - 1))

ci_var_table <- data.frame(
  City  = c("Glasgow", "Edinburgh"),
  Lower = round(c(ci_var_gla[1], ci_var_edi[1]), 7),
  Upper = round(c(ci_var_gla[2], ci_var_edi[2]), 7)
)

knitr::kable(ci_var_table,
             caption = "95\\% confidence intervals for the variance of Attendance.",
             align = "c",
             col.names = c("City", "2.5%", "97.5%"),
             booktabs = TRUE)
```

For the population variance (Table \@ref(tab:ci-var)), the $\chi^2$-based intervals are $(0.00550, 0.00964)$ for Glasgow and $(0.00746, 0.01305)$ for Edinburgh. Unlike the mean intervals, these overlap, suggesting that the difference in variability may not be statistically significant — a question addressed by the formal tests below.

### Hypothesis Test for Means

```{r t-test, echo = FALSE}

t_result <- t.test(glasgow$Attendance, edinburgh$Attendance, var.equal = FALSE)

t_table <- data.frame(
  Statistic = round(t_result$statistic, 4),
  df        = round(t_result$parameter, 2),
  p_value   = formatC(t_result$p.value, format = "e", digits = 4)
)

knitr::kable(t_table,
             caption = "Welch's two-sample t-test for difference in mean Attendance.",
             align = "c",
             col.names = c("t-statistic", "df", "p-value"),
             booktabs = TRUE)
```

To formally assess whether mean attendance differs between the two cities, we test:$$H_0: \mu_{\text{Glasgow}} = \mu_{\text{Edinburgh}} \qquad \text{vs} \qquad H_1: \mu_{\text{Glasgow}} \neq \mu_{\text{Edinburgh}}$$using Welch's two-sample $t$-test. The test yields $t = -2.505$ with $df = 193.61$ and $p = 0.013$. Since $p < \alpha = 0.05$, we reject $H_0$ and conclude that Glasgow's mean attendance rate is statistically significantly lower than Edinburgh's. This finding is consistent with the non-overlapping confidence intervals for the mean reported in Table \@ref(tab:ci-mean).

### Hypothesis Test for Variances

```{r f-test, echo = FALSE}

f_result <- var.test(glasgow$Attendance, edinburgh$Attendance)

f_table <- data.frame(
  Statistic = round(f_result$statistic, 4),
  df1       = f_result$parameter[1],
  df2       = f_result$parameter[2],
  p_value   = formatC(f_result$p.value, format = "e", digits = 4)
)

knitr::kable(f_table,
             caption = "F-test for equality of variances of Attendance.",
             align = "c",
             col.names = c("F-statistic", "df1", "df2", "p-value"),
             booktabs = TRUE)
```

To assess whether attendance rates are more variable in one city than the other, we test:$$H_0: \frac{\sigma^2_{\text{Glasgow}}}{\sigma^2_{\text{Edinburgh}}} = 1 \qquad \text{vs} \qquad H_1: \frac{\sigma^2_{\text{Glasgow}}}{\sigma^2_{\text{Edinburgh}}} \neq 1$$using an $F$-test. The test statistic is $F = s^2_G / s^2_E = 0.00714 / 0.00967 = 0.738$ with $df_1 = df_2 = 99$ and $p = 0.133$. Since $p > \alpha = 0.05$, we fail to reject $H_0$ and conclude that there is insufficient evidence of a significant difference in attendance variability between Glasgow and Edinburgh at the 5% significance level. This is consistent with the overlapping confidence intervals for variance observed in Table \@ref(tab:ci-var).

In summary, Glasgow has a statistically significantly lower mean attendance rate than Edinburgh ($p = 0.013$), while no significant difference in variance is detected ($p = 0.133$). The lower mean attendance in Glasgow, combined with the left-skewed distribution evident in Figure \@ref(fig:city-histogram), suggests that a subset of Glasgow data zones experience particularly poor attendance, which may reflect localised socio-economic disadvantage. These findings motivate the investigation in Section 2, where the analysis explores which socio-economic factors are most strongly associated with attendance across all 400 Scottish data zones.

\newpage
## Section 2: Investigating the Relationship between Attendance and Other Variables

In this section, all 400 data zones are analysed together to evaluate which factors are most closely associated with Attendance. The four variables — Attainment, Crime\_rate, CIF (Comparative Illness Factor), and Income\_rate — were chosen to represent educational attainment, crime, health, and income deprivation, respectively. The linear associations between Attendance and these four variables are compared, and a simple linear regression model is then fitted using the variable with the largest absolute Pearson correlation with Attendance.

### Pairwise Scatterplot Matrix

```{r scatters-1, echo=FALSE, fig.height=6, fig.width=7, fig.cap='Pairwise scatterplot matrix of Attendance and four selected variables'}

vars <- c("Attendance", "Attainment", "crime_rate", "CIF", "Income_rate")

par(family = "serif")
pairs(dat[, vars],
      pch = 16, cex = 0.4, col = adjustcolor(col_point, alpha.f = 0.5),
      main = "",
      labels = c("Attendance", "Attainment", "Crime Rate", "CIF", "Income Rate"))
```

The pairwise scatterplot matrix (Figure \@ref(fig:scatters-1)) suggests that Attendance is positively associated with Attainment and negatively associated with Crime\_rate, CIF and Income\_rate, with the Income\_rate relationship appearing strongest. The matrix also reveals a close association between CIF and Income\_rate, indicating that these two indicators capture closely related aspects of deprivation.

### Pearson Correlation Analysis

```{r Pairwise-Pearson-correlation-coefficients, echo = FALSE}
# Full 5x5 pairwise Pearson correlation matrix
cor_matrix <- round(cor(dat[, vars], use = "complete.obs", method = "pearson"), 3)
colnames(cor_matrix) <- c("Attendance", "Attainment", "Crime Rate", "CIF", "Income Rate")
rownames(cor_matrix) <- c("Attendance", "Attainment", "Crime Rate", "CIF", "Income Rate")

knitr::kable(cor_matrix,
             caption = "Pairwise Pearson correlation coefficients between the five selected variables.",
             align = "c",
             booktabs = TRUE)

# Store individual correlations for later use
r_attainment <- cor_matrix[1, 2]
r_crime      <- cor_matrix[1, 3]
r_cif        <- cor_matrix[1, 4]
r_income     <- cor_matrix[1, 5]
```


To quantify these associations, pairwise Pearson correlation coefficients are computed (Table \@ref(tab:Pairwise-Pearson-correlation-coefficients)). Attendance correlates most strongly with Income\_rate ($r=-0.756$), followed by CIF ($r=-0.735$) and Attainment ($r=0.686$); the weakest association is with Crime\_rate ($r = -0.469$). Income\_rate was therefore selected as the predictor for the simple linear regression model. Notably, the correlation between CIF and Income\_rate is extremely high ($r=0.953$), confirming the substantial overlap noted above.

### Simple Linear Regression Model

```{r linear-regression-model, echo = FALSE}
# Identify the variable with the highest absolute correlation with Attendance
correlations <- c(
  Attainment = r_attainment,
  Crime_rate = r_crime,
  CIF = r_cif,
  Income_rate = r_income
)
data2_name <- names(which.max(abs(correlations)))

# Fit simple linear regression
model_formula <- as.formula(paste("Attendance ~", data2_name))
lm_model <- lm(model_formula, data = dat)

# Extract coefficients
intercept <- coef(lm_model)[1]
slope     <- coef(lm_model)[2]
r_squared <- summary(lm_model)$r.squared
```

We therefore fit a simple linear regression model:

$$\text{Attendance} = \beta_0 + \beta_1\times \text{Income\_Rate} + \varepsilon$$
The estimated regression equation is: 

$$ \hat{y} = 0.8887 + (-0.635) \times x $$
```{r regression-coef-table, echo = FALSE}
display_name <- gsub("_", " ", data2_name)

coef_df <- tidy(lm_model)
coef_df$p.value <- ifelse(coef_df$p.value < 0.001, "< 0.001",
                          formatC(coef_df$p.value, format = "f", digits = 4))
knitr::kable(
  coef_df,
  digits = 4,
  caption = paste("Simple linear regression model: Attendance ~", display_name),
  booktabs = TRUE
)
```


Both estimated coefficients are highly significant ($p<0.001$; Table \@ref(tab:regression-coef-table)). The slope of $-0.6350$ means that a $0.1$ increase in Income\_rate is associated with an average decrease of about $0.0635$ in Attendance — a substantial practical effect.

```{r regression-ci-table, echo = FALSE}
ci <- confint(lm_model, level = 0.95)
rownames(ci) <- c("Intercept ", "Slope")
colnames(ci) <- c("2.5%", "97.5%")

knitr::kable(
  ci,
  digits = 4,
  caption = "95\\% Confidence Intervals for Regression Coefficients",
  booktabs = TRUE
)
```

The $95\%$ confidence interval for the slope (Table \@ref(tab:regression-ci-table)) is entirely negative, reinforcing the clear negative association. With \textbf{$R^2=0.5796$}, approximately $57.96\%$ of the sample variation in Attendance is explained by Income\_rate alone, confirming it as the strongest single linear predictor among the four candidates.

### Fitted Regression Line

```{r linear-regression, echo = FALSE, fig.height=4, fig.width=6, fig.cap='Scatterplot with fitted regression line'}

equation <- paste0("y = ", round(intercept, 3), " + (", round(slope, 3), ") x")

# Dynamic annotation position
x_pos <- quantile(dat[[data2_name]], 0.75, na.rm = TRUE)
y_pos <- quantile(dat$Attendance, 0.95, na.rm = TRUE)

ggplot(dat, aes(x = !!sym(data2_name), y = Attendance)) +
  geom_point(shape = 16, size = 1, colour = adjustcolor(col_point, alpha.f = 0.5)) +
  geom_smooth(method = "lm", se = TRUE, colour = col_line, fill = adjustcolor(col_line, alpha.f = 0.15)) +
  labs(
    x = data2_name,
    y = "Attendance"
  ) +
  annotate("text", x = x_pos, y = y_pos, label = equation, parse = FALSE,
           size = 4, colour = col_line)

```


The negative linear trend is clearly visible in the scatterplot with fitted regression line (Figure \@ref(fig:linear-regression)): data zones with higher income deprivation tend to have lower Attendance rates. However, noticeable vertical scatter persists around the line, particularly at higher values of Income\_rate, suggesting that additional factors also influence Attendance.

### Residual Diagnostics

```{r residuals, echo = FALSE, fig.height=4, fig.width=6, fig.cap='Residuals of the fitted model'}

# residuals
dat$residuals <- residuals(lm_model)
dat$fitted    <- fitted(lm_model)

ggplot(dat, aes(x = fitted, y = residuals)) +
  geom_point(shape = 16, size = 1, colour = adjustcolor(col_point, alpha.f = 0.5)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = col_ref, linewidth = 0.6) +
  labs(
    x = "Predicted Attendance Rate",
    y = "Residuals"
  )

```

The residual plot (Figure \@ref(fig:residuals)) shows no strong curved pattern, supporting the linearity assumption. However, the spread is not perfectly constant — indicating mild heteroscedasticity — and a few relatively large residuals remain, consistent with the observation that factors beyond Income\_rate also influence Attendance.

Overall, among the four candidate variables, Income\_rate shows the strongest negative linear association with Attendance. Therefore, Income\_rate provides the most informative single-variable linear model among the four predictors considered. This finding is consistent with the well-established understanding that poorer economic conditions are associated with lower attendance rates. However, the very high correlation between Income\_rate and CIF indicates substantial overlap between these two deprivation-related variables. In addition, the residual plot shows that some variation in Attendance remains unexplained, so Income\_rate alone cannot fully account for differences in Attendance across data zones.

\newpage

## Section 3: A Further Investigation of the Low Attendance Rate in Glasgow

Section 1 showed that Glasgow has a lower mean Attendance than Edinburgh, while Section 2 identified Income\_rate as the strongest single linear predictor of Attendance in the full sample. The question remains whether Glasgow’s lower Attendance can be explained entirely by income deprivation. To investigate this, we first compare Attendance across Glasgow, Edinburgh and Elsewhere, then compare region-specific regression models, and finally examine Glasgow’s deviation from a non-Glasgow baseline together with its associations with other variables.

### Evaluate Glasgow’s Distinctiveness: Regional Comparison of Attendance

```{r boxplot-3region, echo=FALSE, fig.height=4, fig.width=6, fig.cap='Distribution of Attendance across three regions'}
ggplot(dat, aes(x = Region, y = Attendance, fill = Region)) +
  geom_boxplot(outlier.shape = 1, outlier.size = 1.5) +
  scale_fill_manual(values = c("Glasgow" = col_gla, "Edinburgh" = col_edi, "Elsewhere" = col_else)) +
  labs(x = "Region", y = "Attendance (proportion)") +
  theme(legend.position = "none")
```

As established in Section 1, Glasgow has a lower mean Attendance rate than Edinburgh, with no significant difference in variance. Figure \@ref(fig:boxplot-3region) extends the comparison to Elsewhere in Scotland. Glasgow has the lowest sample mean Attendance $(0.7613)$, compared with $0.7938$ for Edinburgh and $0.8159$ for Elsewhere. This suggests that Glasgow’s low Attendance is not simply a Glasgow–Edinburgh contrast but a broader regional pattern. To confirm this statistically, pairwise F-tests are performed to determine whether variances differ, followed by pairwise Welch's t-tests for means.

```{r f-tests-pairwise, echo=FALSE}
# F-tests: variance comparison
f_gla_else <- var.test(glasgow$Attendance, elsewhere$Attendance)
f_edi_else <- var.test(edinburgh$Attendance, elsewhere$Attendance)
f_gla_edi  <- var.test(glasgow$Attendance, edinburgh$Attendance)

f_table <- data.frame(
  Comparison = c("Glasgow vs Elsewhere", "Edinburgh vs Elsewhere", "Glasgow vs Edinburgh"),
  F_stat     = round(c(f_gla_else$statistic, f_edi_else$statistic, f_gla_edi$statistic), 4),
  df1        = c(f_gla_else$parameter[1], f_edi_else$parameter[1], f_gla_edi$parameter[1]),
  df2        = c(f_gla_else$parameter[2], f_edi_else$parameter[2], f_gla_edi$parameter[2]),
  p_value    = c(formatC(f_gla_else$p.value, format = "e", digits = 3),
                 formatC(f_edi_else$p.value, format = "e", digits = 3),
                 formatC(f_gla_edi$p.value, format = "e", digits = 3))
)

knitr::kable(f_table,
             caption = "F-tests comparing variance of Attendance between regions.",
             align = "c",
             col.names = c("Comparison", "F-statistic", "df1", "df2", "p-value"),
             booktabs = TRUE)
```

```{r t-tests-pairwise, echo=FALSE}
# Welch's t-tests: mean comparison
t_gla_else <- t.test(glasgow$Attendance, elsewhere$Attendance, var.equal = FALSE)
t_edi_else <- t.test(edinburgh$Attendance, elsewhere$Attendance, var.equal = FALSE)
t_gla_edi  <- t.test(glasgow$Attendance, edinburgh$Attendance, var.equal = FALSE)

t_table <- data.frame(
  Comparison = c("Glasgow vs Elsewhere", "Edinburgh vs Elsewhere", "Glasgow vs Edinburgh"),
  t_stat     = round(c(t_gla_else$statistic, t_edi_else$statistic, t_gla_edi$statistic), 4),
  df         = round(c(t_gla_else$parameter, t_edi_else$parameter, t_gla_edi$parameter), 2),
  p_value    = c(formatC(t_gla_else$p.value, format = "e", digits = 3),
                 formatC(t_edi_else$p.value, format = "e", digits = 3),
                 formatC(t_gla_edi$p.value, format = "e", digits = 3))
)

knitr::kable(t_table,
             caption = "Welch's t-tests comparing mean Attendance between regions.",
             align = "c",
             col.names = c("Comparison", "t-statistic", "df", "p-value"),
             booktabs = TRUE)
```

None of the pairwise F-tests (Table \@ref(tab:f-tests-pairwise)) reach significance at the $5\%$ level, so Attendance variability does not differ detectably across regions. The t-tests (Table \@ref(tab:t-tests-pairwise)), however, reveal that Glasgow's mean Attendance is significantly lower than both Edinburgh's ($p=0.0131$) and Elsewhere's ($p=3.784\times10^{-7}$), while Edinburgh and Elsewhere do not differ significantly ($p=0.0572$). Glasgow is thus the distinctively low-Attendance region.



### Income Cannot Fully Explain Low Attendance Rate in Glasgow

If Income\_rate fully explained the Attendance rate in Glasgow, regional regressions should produce similar slopes and $R^2$ values. However, as shown in Table \@ref(tab:regional-regression) and Figure \@ref(fig:scatter-regional), Glasgow's regression line is flatter (slope = $-0.455$ vs $-0.778$ for Elsewhere) and its $R^2$ ($0.3949$) is notably lower, meaning Income\_rate explains less of the variation in Glasgow's attendance.

```{r regional-regression, echo=FALSE}
lm_gla  <- lm(Attendance ~ Income_rate, data = glasgow)
lm_edi  <- lm(Attendance ~ Income_rate, data = edinburgh)
lm_else <- lm(Attendance ~ Income_rate, data = elsewhere)

# Extract coefficients
slopes     <- c(coef(lm_gla)[2], coef(lm_edi)[2], coef(lm_else)[2])
intercepts <- c(coef(lm_gla)[1], coef(lm_edi)[1], coef(lm_else)[1])
r_sq       <- c(summary(lm_gla)$r.squared, summary(lm_edi)$r.squared, summary(lm_else)$r.squared)

# 95% CIs for slopes
ci_gla  <- confint(lm_gla,  "Income_rate", level = 0.95)
ci_edi  <- confint(lm_edi,  "Income_rate", level = 0.95)
ci_else <- confint(lm_else, "Income_rate", level = 0.95)

reg_table <- data.frame(
  Region    = c("Glasgow", "Edinburgh", "Elsewhere"),
  Intercept = round(intercepts, 4),
  Slope     = round(slopes, 4),
  CI_lower  = round(c(ci_gla[1], ci_edi[1], ci_else[1]), 4),
  CI_upper  = round(c(ci_gla[2], ci_edi[2], ci_else[2]), 4),
  R_squared = round(r_sq, 4)
)

knitr::kable(reg_table,
             caption = "Regional regression models: Attendance on Income\\_rate.",
             align = "c",
             booktabs = TRUE,
             col.names = c("Region", "Intercept", "Slope",
                           "Slope CI lower", "Slope CI upper", "R-squared"))
```

```{r scatter-regional, echo=FALSE, fig.height=4, fig.width=6, fig.cap='Regional regression lines for Attendance on Income\\_rate'}
ggplot(dat, aes(x = Income_rate, y = Attendance, colour = Region)) +
  geom_point(shape = 16, size = 1, alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.15) +
  scale_colour_manual(values = c("Glasgow" = col_gla, "Edinburgh" = col_edi, "Elsewhere" = col_else)) +
  labs(x = "Income_rate", y = "Attendance rate")
```

```{r prediction-intervals, echo=FALSE}
new_data <- data.frame(Income_rate = 0.15)

pred_gla  <- predict(lm_gla,  newdata = new_data, interval = "prediction", level = 0.95)
pred_edi  <- predict(lm_edi,  newdata = new_data, interval = "prediction", level = 0.95)
pred_else <- predict(lm_else, newdata = new_data, interval = "prediction", level = 0.95)

pred_table <- data.frame(
  Region = c("Glasgow", "Edinburgh", "Elsewhere"),
  Fitted = round(c(pred_gla[1], pred_edi[1], pred_else[1]), 4),
  Lower  = round(c(pred_gla[2], pred_edi[2], pred_else[2]), 4),
  Upper  = round(c(pred_gla[3], pred_edi[3], pred_else[3]), 4),
  Width  = round(c(pred_gla[3] - pred_gla[2],
                    pred_edi[3] - pred_edi[2],
                    pred_else[3] - pred_else[2]), 4)
)

knitr::kable(pred_table,
             caption = "95\\% prediction intervals for Attendance at Income\\_rate $= 0.15$.",
             align = "c",
             col.names = c("Region", "Fitted", "Lower", "Upper", "Width"),
             booktabs = TRUE)
```

To further illustrate this, Table \@ref(tab:prediction-intervals) compares Attendance predictions at a common Income\_rate level ($= 0.15$, close to the overall median). Glasgow's predicted Attendance is lower and its prediction interval is notably wider, reflecting greater unexplained variability. Combined with Glasgow's much lower $R^2$ and flatter slope, this demonstrates that the income--attendance relationship operates fundamentally differently in Glasgow --- income deprivation alone cannot fully account for Glasgow's attendance patterns, suggesting the existence of a \textbf{"Glasgow effect"}.

### Quantifying the “Glasgow Effect”: Baseline Model and Residual Analysis

```{r baseline-model, echo=FALSE}
# Baseline model: trained on non-Glasgow data
dat_baseline <- subset(dat, Region != "Glasgow")
lm_baseline  <- lm(Attendance ~ Income_rate, data = dat_baseline)

baseline_summary <- summary(lm_baseline)
b0 <- coef(lm_baseline)[1]
b1 <- coef(lm_baseline)[2]
```

To assess Glasgow’s deviation from the non-Glasgow pattern, a baseline regression model is fitted based on the non-Glasgow observations, as shown in Table \@ref(tab:baseline-table). For each Glasgow data zone, we then compare the observed Attendance with the value predicted by this baseline model. Figure \@ref(fig:residual-histogram) shows the distribution of these residual deviations.

```{r baseline-table, echo=FALSE}
baseline_ci <- confint(lm_baseline, level = 0.95)

baseline_coef <- data.frame(
  Term      = c("Intercept", "Income_rate"),
  Estimate  = round(coef(lm_baseline), 4),
  CI_lower  = round(baseline_ci[, 1], 4),
  CI_upper  = round(baseline_ci[, 2], 4)
)

knitr::kable(baseline_coef,
             caption = paste0("Baseline regression model (non-Glasgow, $n = ",
                              nrow(dat_baseline), "$): Attendance on Income\\_rate."),
             align = "c",
             booktabs = TRUE,
             col.names = c("Term", "Estimate", "95% CI lower", "95% CI upper"),
             row.names = FALSE)
```

```{r glasgow-residuals, echo=FALSE}
# Predict Glasgow's "expected" attendance using the baseline model
glasgow$predicted <- predict(lm_baseline, newdata = glasgow)
glasgow$residual  <- glasgow$Attendance - glasgow$predicted

# Descriptive statistics of Glasgow residuals
mean_resid <- mean(glasgow$residual)
sd_resid   <- sd(glasgow$residual)
n_resid    <- length(glasgow$residual)
```

```{r residual-histogram, echo=FALSE, fig.height=4, fig.width=6, fig.cap='Distribution of Glasgow residuals (Actual $-$ Predicted Attendance)'}
resid_df <- data.frame(residual = glasgow$residual)
ggplot(resid_df, aes(x = residual)) +
  geom_histogram(bins = 15, fill = col_gla, colour = "white", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = col_ref, linewidth = 0.6) +
  geom_vline(xintercept = mean_resid, colour = col_line, linewidth = 0.8) +
  annotate("text", x = mean_resid, y = Inf, vjust = 2, hjust = -0.1,
           label = paste0("Mean = ", round(mean_resid, 4)),
           colour = col_line, size = 3.5) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 6)) +
  labs(x = "Residual (Actual - Predicted)", y = "Frequency") +
  theme(
        axis.text.x = element_text(size = 9),
        plot.margin = margin(t = 5, r = 10, b = 15, l = 5, unit = "pt"))
```

To test whether the mean deviation differs significantly from zero, a one-sample t-test is performed with hypotheses:
$$H_0: \mu_e = 0 \quad \text{(no Glasgow effect)} \qquad H_1: \mu_e \neq 0$$

```{r residual-ttest, echo=FALSE}
t_resid <- t.test(glasgow$residual, mu = 0)

t_resid_table <- data.frame(
  Mean_residual = round(mean_resid, 4),
  SE            = round(sd_resid / sqrt(n_resid), 4),
  t_stat        = round(t_resid$statistic, 4),
  df            = t_resid$parameter,
  p_value       = formatC(t_resid$p.value, format = "e", digits = 3),
  CI_lower      = round(t_resid$conf.int[1], 4),
  CI_upper      = round(t_resid$conf.int[2], 4)
)

knitr::kable(t_resid_table,
             caption = "One-sample t-test on Glasgow residuals ($H_0: \\mu_e = 0$).",
             align = "c",
             booktabs = TRUE,
             col.names = c("Mean residual", "SE", "t-statistic", "df",
                           "p-value", "95% CI lower", "95% CI upper"),
             row.names = FALSE)
```

The $p$-value is `r formatC(t_resid$p.value, format = "e", digits = 3)` and the $95\%$ confidence interval for the mean residual excludes zero. The mean residual is $\bar{e} =$ `r round(mean_resid, 4)` $> 0$, which may seem surprising given that Glasgow has the lowest raw attendance among the three regions. However, Glasgow also has by far the highest income deprivation, so the non-Glasgow baseline predicts very low attendance for Glasgow's income levels. The positive residual indicates that, conditional on its extreme income deprivation, Glasgow's attendance is actually somewhat higher than the non-Glasgow trend would predict. Therefore, the \textbf{"Glasgow effect"} here means a fundamentally different income--attendance relationship --- as evidenced by Glasgow's flatter slope and lower $R^2$ ($\approx 0.39$) in the regional regression analysis --- not simply an additional negative penalty on top of income deprivation.

### Identifying the Source: Residual Correlation Scan

To explore which additional factors are associated with Glasgow’s deviation from the non-Glasgow baseline, Pearson correlations were computed between the Glasgow residuals and candidate variables.

```{r residual-scan, echo=FALSE}
# Select continuous candidate variables (exclude identifiers, Income, and Attendance)
scan_vars <- c("Employment_rate", "CIF",
               "ALCOHOL", "DRUG", "SMR", "DEPRESS", "LBWT", "EMERG",
               "Attainment", "no_qualifications", "not_participating", "University",
               "drive_petrol", "drive_GP", "drive_post", "drive_primary",
               "drive_retail", "drive_secondary",
               "PT_GP", "PT_post", "PT_retail",
               "Broadband", "crime_rate",
               "overcrowded_rate", "nocentralheat_rate")

# Compute Pearson r between Glasgow residuals and each variable
cors <- sapply(scan_vars, function(v) {
  x <- glasgow[[v]]
  if (is.numeric(x) && sum(!is.na(x)) > 5) {
    cor(glasgow$residual, x, use = "complete.obs")
  } else {
    NA
  }
})

# Sort by absolute correlation
cors <- cors[!is.na(cors)]
cors_sorted <- sort(abs(cors), decreasing = TRUE)
cors_signed <- cors[names(cors_sorted)]

scan_table <- data.frame(
  Variable = names(cors_signed),
  r        = round(cors_signed, 4),
  abs_r    = round(abs(cors_signed), 4)
)

knitr::kable(scan_table,
             caption = "Pearson correlations: Glasgow residuals vs candidate variables (sorted by absolute r).",
             align = "c",
             col.names = c("Variable", "r", "Abs(r)"),
             row.names = FALSE,
             booktabs = TRUE)
```

```{r identify-top-var, echo=FALSE}
# Identify the variable with the highest |r|
top_var_name <- names(cors_signed)[1]
top_var_r    <- cors_signed[1]
top_var_tex  <- gsub("_", "\\_", top_var_name, fixed = TRUE)
```

Among all candidate variables (Table \@ref(tab:residual-scan)), `r top_var_tex` shows the strongest correlation with the Glasgow residuals ($r =$ `r round(top_var_r, 3)`), suggesting it is the leading candidate for explaining Glasgow's deviation from the non-Glasgow baseline after accounting for income deprivation. This scan is exploratory: it identifies associated variables but does not establish causality, and some candidates may overlap with broader deprivation conditions.

### Deep Dive: Further Analysis of the Leading Candidate Variable

```{r deep-dive-scatter, echo=FALSE, fig.height=4, fig.width=6, fig.cap=paste0('Glasgow residuals vs ', top_var_tex)}
lm_deep <- lm(glasgow$residual ~ glasgow[[top_var_name]])
r_deep     <- cor(glasgow$residual, glasgow[[top_var_name]], use = "complete.obs")
r2_deep    <- summary(lm_deep)$r.squared
slope_deep <- coef(lm_deep)[2]
ci_deep    <- confint(lm_deep, level = 0.95)

deep_df <- data.frame(x = glasgow[[top_var_name]], y = glasgow$residual)

ggplot(deep_df, aes(x = x, y = y)) +
  geom_point(shape = 16, size = 1, alpha = 0.4, colour = col_gla) +
  geom_smooth(method = "lm", se = TRUE, colour = col_line,
              fill = adjustcolor(col_line, alpha.f = 0.15), linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = col_ref, linewidth = 0.6) +
  labs(x = top_var_name, y = "Residual (Actual - Predicted Attendance)")
```

```{r deep-dive-table, echo=FALSE}
deep_table <- data.frame(
  Statistic = c("Pearson r", "R-squared", "Slope", "Slope 95% CI lower", "Slope 95% CI upper"),
  Value     = round(c(r_deep, r2_deep, slope_deep, ci_deep[2, 1], ci_deep[2, 2]), 4)
)

knitr::kable(deep_table,
             caption = paste0("Regression of Glasgow residuals on ", top_var_tex, "."),
             align = "c",
             row.names = FALSE,
             booktabs = TRUE)
```

Within Glasgow, data zones with higher `r top_var_tex` — i.e.\ a greater percentage of employment-deprived residents — tend to have more positive residuals (Figure \@ref(fig:deep-dive-scatter); Table \@ref(tab:deep-dive-table)), meaning their attendance exceeds the income-only baseline prediction. This does not imply that employment deprivation improves attendance; rather, the income-based model under-predicts attendance in the most employment-deprived Glasgow areas, pointing to unmeasured offsetting factors. The regression of residuals on `r top_var_tex` explains $R^2 =$ `r round(r2_deep, 3)` of the variation in the Glasgow effect, with a slope whose $95\%$ confidence interval excludes zero.

```{r chisq-bonus, echo=FALSE}
q_resid <- quantile(glasgow$residual, probs = c(1/3, 2/3))
q_topvar <- quantile(glasgow[[top_var_name]], probs = c(1/3, 2/3), na.rm = TRUE)

resid_cat <- cut(glasgow$residual,
                 breaks = c(-Inf, q_resid[1], q_resid[2], Inf),
                 labels = c("Negative", "Neutral", "Positive"))
topvar_cat <- cut(glasgow[[top_var_name]],
                  breaks = c(-Inf, q_topvar[1], q_topvar[2], Inf),
                  labels = c("Low", "Medium", "High"))

cont_table <- table(resid_cat, topvar_cat)

knitr::kable(cont_table,
             caption = paste0("Contingency table: Residual category $\\times$ ",
                              top_var_tex, " category (Glasgow only)."),
             booktabs = TRUE)

chi_result <- chisq.test(cont_table)
```

As a cross-validation using a different statistical framework, a $\chi^2$ test of independence is performed. Categorising both the residuals and `r top_var_tex` into terciles yields the contingency table above. The test statistic is $\chi^2 =$ `r round(chi_result$statistic, 3)` with `r chi_result$parameter` degrees of freedom ($p =$ `r formatC(chi_result$p.value, format = "e", digits = 3)`). `r ifelse(chi_result$p.value < 0.05, "We reject $H_0$ of independence: the two variables are significantly associated, confirming the linear analysis.", "The result is not significant at the 5% level, suggesting the association is primarily captured by the linear relationship.")`

### Section 3 Summary

This investigation followed a four-stage analytical arc:

1. **Glasgow has distinctively lower attendance.** Pairwise tests confirm that Glasgow's mean attendance is significantly lower than both Edinburgh's ($p = 0.013$) and Elsewhere's ($p < 0.001$), while Edinburgh and Elsewhere do not differ significantly ($p = 0.057$) — the deficit is Glasgow-specific.
2. **Income cannot fully explain it.** Separate regional regressions show that Glasgow has a flatter slope and lower $R^2$ (`r round(r_sq[1], 3)`) compared to Edinburgh (`r round(r_sq[2], 3)`) and Elsewhere (`r round(r_sq[3], 3)`). At the same income level, Glasgow's prediction interval is markedly wider.
3. **The "Glasgow effect" is real but nuanced.** A baseline model trained on non-Glasgow data yields a positive mean residual for Glasgow ($\bar{e} =$ `r round(mean_resid, 4)`, $p =$ `r formatC(t_resid$p.value, format = "e", digits = 3)`): given its extreme income deprivation, Glasgow's attendance is actually higher than the non-Glasgow trend would predict. The "Glasgow effect" therefore reflects a fundamentally different income--attendance relationship, not an additional negative penalty.
4. **`r top_var_tex` is the leading candidate explanation.** A correlation scan of the residuals identifies `r top_var_tex` ($r =$ `r round(top_var_r, 3)`) as the variable most strongly associated with the Glasgow effect, confirmed by both regression analysis and a $\chi^2$ independence test.

These findings suggest that addressing Glasgow's attendance gap may require interventions beyond poverty reduction. Factors related to `r top_var_tex` appear to play a significant additional role and merit further investigation.

\newpage


# Conclusions

This report investigated the distribution of school attendance across Scottish data zones, its association with socio-economic deprivation indicators, and whether income deprivation alone can account for Glasgow's distinctively low attendance. 

In the first section, we compared Attendance between Glasgow and Edinburgh. The results showed that Glasgow has a lower mean Attendance rate than Edinburgh. The non-overlapping $95\%$ confidence intervals for the means and the significant Welch's $t$-test both indicate that mean Attendance is lower in Glasgow than in Edinburgh. However, the confidence intervals for the variances overlapped, and the F-test found no statistically significant evidence that the variability of Attendance differs between Glasgow and Edinburgh.

In the second section, we examined the relationship between Attendance and four selected variables using the full sample of 400 data zones. The pairwise scatterplot matrix and Pearson correlation coefficients showed that Income\_rate had the strongest absolute correlation with Attendance among the four selected variables (Attainment, Crime\_rate, CIF, and Income\_rate). The fitted simple linear regression model confirmed a clear negative relationship between Income\_rate and Attendance: data zones with higher income deprivation tend to have lower Attendance rates. The model explained approximately $57.96\%$ of the variation in Attendance. The residual diagnostics suggested that the linear model was broadly reasonable, although some unexplained variation remained.

In the third section, we investigated whether Glasgow's low Attendance can be explained entirely by income deprivation. Pairwise comparisons confirmed that Glasgow is the distinctive low-Attendance region, with significantly lower mean Attendance than both Edinburgh and Elsewhere, while Edinburgh and Elsewhere did not differ significantly. Region-specific regressions revealed that Income\_rate explains less of the variation in Glasgow's attendance ($R^2 \approx 0.39$) than in Edinburgh or Elsewhere, with a flatter slope and wider prediction intervals. A baseline model trained on non-Glasgow data showed that Glasgow's attendance deviates significantly from the non-Glasgow baseline; notably, the mean residual is positive, indicating that given Glasgow's extreme income deprivation, its attendance is actually higher than the non-Glasgow trend would predict. The "Glasgow effect" therefore reflects a fundamentally different income--attendance relationship rather than a simple additional penalty. A residual correlation scan identified `r top_var_tex` as the variable most strongly associated with this residual pattern, and the association was further supported by regression analysis and a $\chi^2$ independence test. These findings suggest that addressing Glasgow's attendance gap may require interventions beyond poverty reduction alone.

There are several limitations to this analysis. First, the dataset is cross-sectional, so all associations reported are correlational rather than causal. Second, the simple linear regression models include only a single predictor; multiple regression could provide a more nuanced picture but was beyond the scope of this report. Third, the chi-squared test in Section 3 required discretising continuous variables into terciles, which inevitably involves some loss of information. Finally, the residual correlation scan is exploratory, and the top-ranked variable may itself be a proxy for other unmeasured confounders.

Future work could extend the analysis in several directions. A multiple regression framework incorporating several predictors simultaneously would help to disentangle the joint effects of income, employment, and other deprivation indicators on attendance. Longitudinal data, if available, would allow investigation of whether changes in socio-economic conditions over time lead to corresponding changes in attendance. Additionally, spatial modelling techniques could account for geographic clustering of data zones, potentially revealing neighbourhood effects that the current analysis cannot capture.

\newpage


# Appendix

The R code to produce the analysis and plots in this report is as follows:

```{r ref.label=knitr::all_labels(), echo = T, eval = F, tidy=TRUE, tidy.opts=list(width.cutoff=70)}
```