---
title: "An analysis of data on ... in Scotland*"
author: "Yuchen Xu, Yu Li, Lingke Huang, Sirui Lu"
date: ""
output:
  bookdown::pdf_document2:
    latex_engine: pdflatex
    number_sections: false
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
dat <- read.csv("scottishData.csv")
library(ggplot2)
library(dplyr)
library(broom)

# Subsets for Section 1
glasgow   <- subset(dat, Council_area == "Glasgow City")
edinburgh <- subset(dat, Council_area == "City of Edinburgh")

# Region variable for Section 3
dat$Region <- ifelse(dat$Council_area == "Glasgow City", "Glasgow",
              ifelse(dat$Council_area == "City of Edinburgh", "Edinburgh", "Elsewhere"))
dat$Region <- factor(dat$Region, levels = c("Glasgow", "Edinburgh", "Elsewhere"))

# Unified colour palette 
col_gla   <- "#D6604D"   # terracotta  — Glasgow
col_edi   <- "#4393C3"   # steel blue  — Edinburgh
col_else  <- "#66C2A5"   # mint green  — Elsewhere
col_point <- "#525252"   # slate gray  — scatter points
col_line  <- "#2166AC"   # deep blue   — regression line
col_ref   <- "#878787"   # medium gray — reference line
```

\newpage
## Summary
This report investigates the distribution of school attendance rates across Scottish data zones and their relationship with socio-economic factors. Utilizing a sample of $N = 400$ data zones drawn from the 2020 Scottish Index of Multiple Deprivation dataset.

The report uses methods of summary statistics with descriptive plots, along with inferential statistics, including hypothesis test and linear regression. To compare attendance metrics between Glasgow City and the City of Edinburgh, Welch's two-sample $t$-test and an $F$-test for equality of variances were conducted. 

As a result, the findings indicate that Glasgow City exhibits a significantly lower mean attendance rate compared to the City of Edinburgh, although there is no statistically significant difference in variance between the two municipalities. Furthermore, correlation analysis and simple linear regression applied to the full sample identify the income deprivation rate as the strongest linear predictor of attendance $r$. Finally, the fitted linear model indicates that there is a clear negative impact of poverty on school attendance.
\newpage


## Introduction
School attendance is a fundamental indicator of educational engagement. Research consistently shows that pupils with high rates of absenteeism achieve lower academic outcomes, face greater difficulties in securing employment, and are more likely to experience social exclusion in later life. In Scotland, school attendance varies considerably across geographic areas, and this variation is thought to reflect underlying differences in socio-economic conditions rather than individual choice alone. Understanding where attendance is lowest, and which factors are most strongly associated with poor attendance, is therefore of direct importance to education authorities and government policymakers seeking to allocate resources effectively.

For the aim of this report, first this report compares the distribution of school pupil attendance rates between Glasgow City and the City of Edinburgh — Scotland's two largest cities — to determine whether the two cities differ significantly in their average attendance levels and in the variability of attendance across their constituent areas. Second, using the full available dataset, this report investigates which socio-economic indicators are most strongly linearly associated with attendance, and we quantify that relationship through a simple linear regression model.

The data used in this report are drawn from the 2020 Scottish Index of Multiple Deprivation (SIMD), a composite measure developed by the Scottish Government to identify concentrations of deprivation across Scotland and to support evidence-based policy-making. The dataset contains observations from a sample of $n = 400$ data zones — small, spatially defined geographic units each typically containing between 500 and 1,000 residents, of which there are 6,976 in Scotland in total. Of the 400 sampled zones, 100 are located in Glasgow City, 100 in the City of Edinburgh, and the remaining 200 are from elsewhere in Scotland. Each data zone is described by 34 variables spanning domains including population, income, employment, health, education, geographic access to services, crime, and housing. The primary response variable of interest is Attendance, defined as the proportion of school-age pupils attending school within a given data zone over the academic year. As a continuous numerical variable taking values in the interval $[0, 1]$, the attendance data in the sample exhibits considerable regional variability, capturing every possible attendant situation from severe absenteeism to those with full attendance records.

This research is important because identifying the specific socio-economic drivers of low attendance allows policymakers to precisely target resources and interventions to the most deprived communities, ultimately helping to bridge the educational inequality gap.
\newpage

## Methods
Firstly, using the methods of comparative Analysis of Glasgow and Edinburgh. To compare the attendance rates between the City of Glasgow ($n_1 = 100$) and the City of Edinburgh ($n_2 = 100$), initially employed exploratory graphical techniques. Side-by-side boxplots were constructed to visually assess differences in medians and to identify potential outliers, while histograms were generated to examine the underlying shape and skewness of each city's attendance distribution.

Following the visual inspection, sample means ($\bar{x}$) and sample variances ($s^2$) were calculated. A $95\%$ confidence interval for each city's population mean was constructed using the $t$-distribution:
$$\bar{x} \pm t_{\alpha/2,\; n-1} \cdot \frac{s}{\sqrt{n}}$$
and a $95\%$ confidence interval for each city's population variance was constructed using the $\chi^2$-distribution:
$$\left(\frac{(n-1)s^2}{\chi^2_{1-\alpha/2,\; n-1}},\quad \frac{(n-1)s^2}{\chi^2_{\alpha/2,\; n-1}}\right)$$
To formally test for a difference in the population mean attendance rates, a Welch's two-sample $t$-test was conducted. This test was selected as it does not assume equal population variances. The hypotheses tested were $H_0: \mu_{\text{Glasgow}} = \mu_{\text{Edinburgh}}$ against the two-sided alternative $H_1: \mu_{\text{Glasgow}} \neq \mu_{\text{Edinburgh}}$. Subsequently, to compare the variability between the two cities, an $F$-test was applied to test the ratio of the population variances, evaluating $H_0:\sigma^2_{\text{Glasgow}} / \sigma^2_{\text{Edinburgh}} = 1$ versus $H_1: \sigma^2_{\text{Glasgow}} / \sigma^2_{\text{Edinburgh}} \neq 1$.


Secondly, correlation and Linear Regression Analysis. For the second stage of the analysis, all $N = 400$ sampled data zones were utilised. To explore the broader relationships between the response variable (Attendance, $y$) and four potential socio-economic predictors (Attainment, Crime Rate, Community Integration Factor, and Income Rate), a pairwise scatterplot matrix was generated. The strength and direction of the linear associations between these variables were then quantified using the Pearson product-moment correlation coefficient:
$$r = \frac{\sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum_{i=1}^{n} (x_i - \bar{x})^2 \sum_{i=1}^{n} (y_i - \bar{y})^2}}$$
Upon identifying the variable with the highest absolute value of $r$ with Attendance as the predictor, a simple linear regression (SLR) model was estimated using Ordinary Least Squares (OLS). The model is specified as:
$$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i, \qquad \epsilon_i \overset{\text{iid}}{\sim} N(0, \sigma^2)$$
where $\beta_0$ is the intercept and $\beta_1$ is the slope coefficient. The overall goodness-of-fit was evaluated using the coefficient of determination $R^2$. Finally, a plot of the residuals ($\hat{\epsilon}_i$) against the fitted values ($\hat{Y}_i$) was examined to verify the assumptions of linearity and homoscedasticity.



\newpage


## Results

### Section 1:Summary Statistics and Hypothesis Test
This section compares the distribution of Attendance between Glasgow City ($n_1 = 100$) and the City of Edinburgh ($n_2 = 100$) using graphical summaries, descriptive statistics, confidence intervals, and formal hypothesis tests.

### 1.1 Exploratory Visualisation

```{r city-boxplot, fig.height=4, fig.width=6, fig.cap='Side-by-side boxplots of Attendance for Glasgow and Edinburgh'}

dat_two <- subset(dat, Council_area %in% c("Glasgow City", "City of Edinburgh"))

ggplot(dat_two, aes(x = Council_area, y = Attendance, fill = Council_area)) +
  geom_boxplot(outlier.shape = 1, outlier.size = 1.5) +
  scale_fill_manual(values = c("City of Edinburgh" = col_edi, "Glasgow City" = col_gla)) +
  labs(x = "City", y = "Attendance", title = "Attendance by City") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
```

Figure \@ref(fig:city-boxplot)  presents side-by-side boxplots of Attendance for the two cities. Edinburgh has a noticeably higher median attendance rate than Glasgow, and Glasgow's distribution displays greater spread, with at least one low-lying outlier indicating that certain Glasgow data zones have particularly poor attendance. 


```{r city-histogram, fig.height=4, fig.width=6, fig.cap='Histograms of Attendance for Glasgow and Edinburgh'}

ggplot(dat_two, aes(x = Attendance, fill = Council_area)) +
  geom_histogram(bins = 20, alpha = 0.7, color = "white") +
  scale_fill_manual(values = c("City of Edinburgh" = col_edi, "Glasgow City" = col_gla)) +
  facet_wrap(~ Council_area, ncol = 2) +
  labs(x = "Attendance", y = "Frequency", title = "Distribution of Attendance by City") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
```

Figure \@ref(fig:city-histogram) shows city-level histograms, which reinforce this impression: Edinburgh's distribution is tightly concentrated at higher values, whereas Glasgow's is wider and exhibits a noticeable left skew, suggesting that a subset of Glasgow data zones have substantially below-average attendance rates.

### 1.2 Descriptive Statistics

```{r mean-var}

n_gla <- sum(!is.na(glasgow$Attendance))
n_edi <- sum(!is.na(edinburgh$Attendance))
mean_gla <- mean(glasgow$Attendance, na.rm = TRUE)
mean_edi <- mean(edinburgh$Attendance, na.rm = TRUE)
var_gla  <- var(glasgow$Attendance, na.rm = TRUE)
var_edi  <- var(edinburgh$Attendance, na.rm = TRUE)

sd_gla   <- sd(glasgow$Attendance, na.rm = TRUE)
sd_edi   <- sd(edinburgh$Attendance, na.rm = TRUE)

stats_table <- data.frame(
  City     = c("Glasgow", "Edinburgh"),
  N        = c(n_gla, n_edi),
  Mean     = round(c(mean_gla, mean_edi), 5),
  SD       = round(c(sd_gla, sd_edi), 5),
  Variance = round(c(var_gla, var_edi), 7)
)

knitr::kable(stats_table,
             caption = "Sample means and variances of Attendance for Glasgow and Edinburgh.",
             align = "c")
```

Table \@ref(tab:mean-var) reports the sample means and variances for each city. Glasgow's mean attendance is $\bar{x}_G = 0.7613$, compared to $\bar{x}_E = 0.7938$ for Edinburgh, a difference of approximately 3.2 percentage points. Glasgow's sample variance is $s^2_G = 0.00714$, which is smaller than Edinburgh's $s^2_E = 0.00967$, suggesting that attendance rates in Glasgow are slightly less dispersed in raw terms, though the visual evidence from \@ref(fig:city-histogram) shows Glasgow's distribution has a heavier lower tail.


### 1.3 Confidence Intervals
```{r ci-mean}

ci_mean_gla <- mean_gla + c(-1, 1) * qt(0.975, df = n_gla - 1) * sqrt(var_gla / n_gla)
ci_mean_edi <- mean_edi + c(-1, 1) * qt(0.975, df = n_edi - 1) * sqrt(var_edi / n_edi)

ci_mean_table <- data.frame(
  City  = c("Glasgow", "Edinburgh"),
  Lower = round(c(ci_mean_gla[1], ci_mean_edi[1]), 5),
  Upper = round(c(ci_mean_gla[2], ci_mean_edi[2]), 5)
)

knitr::kable(ci_mean_table,
             caption = "95\\% confidence intervals for the mean of Attendance.",
             align = "c",
             col.names = c("City", "2.5%", "97.5%"))
```

Table \@ref(tab:ci-mean) presents the $95\%$ confidence intervals for the population mean attendance in each city. Glasgow's interval is $(0.745, 0.778)$ and Edinburgh's is $(0.774, 0.813)$. These two intervals do not overlap, providing strong preliminary evidence that the true mean attendance rates differ significantly between the two cities. 

```{r ci-var, echo = FALSE}

# 95% CI for variance using chi-squared distribution
ci_var_gla <- c((n_gla - 1) * var_gla / qchisq(0.975, df = n_gla - 1),
                (n_gla - 1) * var_gla / qchisq(0.025, df = n_gla - 1))
ci_var_edi <- c((n_edi - 1) * var_edi / qchisq(0.975, df = n_edi - 1),
                (n_edi - 1) * var_edi / qchisq(0.025, df = n_edi - 1))

ci_var_table <- data.frame(
  City  = c("Glasgow", "Edinburgh"),
  Lower = round(c(ci_var_gla[1], ci_var_edi[1]), 7),
  Upper = round(c(ci_var_gla[2], ci_var_edi[2]), 7)
)

knitr::kable(ci_var_table,
             caption = "95\\% confidence intervals for the variance of Attendance.",
             align = "c",
             col.names = c("City", "2.5%", "97.5%"))
```

Table \@ref(tab:ci-var) presents the $95\%$ confidence intervals for the population variance, constructed using the $\chi^2$-distribution. Glasgow's interval is $(0.00550, 0.00964)$ and Edinburgh's is $(0.00746, 0.01305)$. Unlike the mean intervals, these two intervals overlap, suggesting that the difference in variability between the two cities may not be statistically significant; this is investigated formally below.

### 1.4 Test for Difference in Means

```{r t-test, echo = FALSE}

t_result <- t.test(glasgow$Attendance, edinburgh$Attendance, var.equal = FALSE)

t_table <- data.frame(
  Statistic = round(t_result$statistic, 4),
  df        = round(t_result$parameter, 2),
  p_value   = formatC(t_result$p.value, format = "e", digits = 4)
)

knitr::kable(t_table,
             caption = "Welch's two-sample t-test for difference in mean Attendance.",
             align = "c",
             col.names = c("t-statistic", "df", "p-value"))
```

To formally assess whether mean attendance differs between the two cities, we test:$$H_0: \mu_{\text{Glasgow}} = \mu_{\text{Edinburgh}} \qquad \text{vs} \qquad H_1: \mu_{\text{Glasgow}} \neq \mu_{\text{Edinburgh}}$$using Welch's two-sample $t$-test. The test yields $t = -2.505$ with $df = 193.61$ and $p = 0.013$. Since $p < \alpha = 0.05$, we reject $H_0$ and conclude that Glasgow's mean attendance rate is statistically significantly lower than Edinburgh's. This finding is consistent with the non-overlapping confidence intervals for the mean reported in Table \@ref(tab:ci-mean).

### 1.5 Test for Difference in Variability

```{r f-test, echo = FALSE}

f_result <- var.test(glasgow$Attendance, edinburgh$Attendance)

f_table <- data.frame(
  Statistic = round(f_result$statistic, 4),
  df1       = f_result$parameter[1],
  df2       = f_result$parameter[2],
  p_value   = formatC(f_result$p.value, format = "e", digits = 4)
)

knitr::kable(f_table,
             caption = "F-test for equality of variances of Attendance.",
             align = "c",
             col.names = c("F-statistic", "df1", "df2", "p-value"))
```

To assess whether attendance rates are more variable in one city than the other, we test:$$H_0: \frac{\sigma^2_{\text{Glasgow}}}{\sigma^2_{\text{Edinburgh}}} = 1 \qquad \text{vs} \qquad H_1: \frac{\sigma^2_{\text{Glasgow}}}{\sigma^2_{\text{Edinburgh}}} \neq 1$$using an $F$-test. The test statistic is $F = s^2_G / s^2_E = 0.00714 / 0.00967 = 0.738$ with $df_1 = df_2 = 99$ and $p = 0.133$. Since $p > \alpha = 0.05$, we fail to reject $H_0$ and conclude that there is insufficient evidence of a significant difference in attendance variability between Glasgow and Edinburgh at the $5\%$ significance level. This is consistent with the overlapping confidence intervals for variance observed in Table \@ref(tab:ci-var).

\newpage
### Section 2: Investigating the relationship between Attendance and other variables

In this section, all 400 data zones are used to evaluate which factors are most closely associated with Attendance, without distinguishing whether they are from  Glasgow, Edinburgh, or elsewhere in Scotland. The Attainment, Crime\_rate, CIF (Community Integration Factor), and Income\_rate are selected, because they represent different socio-economic and educational dimensions that may be associated with school attendance.

### 2.1 Pairwise Relationships

```{r scatters-1, echo=FALSE, fig.height=6, fig.width=7, fig.cap='Pairwise scatterplot matrix of Attendance and four selected variables'}

vars <- c("Attendance", "Attainment", "crime_rate", "CIF", "Income_rate")

pairs(dat[, vars],
      pch = 16, cex = 0.4, col = adjustcolor(col_point, alpha.f = 0.5),
      main = "",
      labels = c("Attendance", "Attainment", "Crime Rate", "CIF", "Income Rate"))
```

Figure \@ref(fig:scatters-1) shows the pairwise scatterplot matrix for Attendance and the four selected variables. It suggests approximately linear negative associations between Attendance and both Income\_rate and CIF, a moderate negative association with Crime\_rate, and a positive association with Attainment. Attendance has positive association with Attainment, and negative relationships with Crime\_rate, CIF and Income\_rate. Among these relationships, the strongest association is between Attendance and Income\_rate. In addition, there is a very strongly positively association between CIF and Income\_rate, indicating overlap of deprivations. 

### 2.2 Pearson Correlation Coefficients

```{r Pairwise-Pearson-correlation-coefficients, echo = FALSE}
# Full 5x5 pairwise Pearson correlation matrix
cor_matrix <- round(cor(dat[, vars], use = "complete.obs", method = "pearson"), 3)
colnames(cor_matrix) <- c("Attendance", "Attainment", "Crime Rate", "CIF", "Income Rate")
rownames(cor_matrix) <- c("Attendance", "Attainment", "Crime Rate", "CIF", "Income Rate")

knitr::kable(cor_matrix,
             caption = "Pairwise Pearson correlation coefficients between the five selected variables.",
             align = "c")

# Store individual correlations for later use
r_attainment <- cor_matrix[1, 2]
r_crime      <- cor_matrix[1, 3]
r_cif        <- cor_matrix[1, 4]
r_income     <- cor_matrix[1, 5]
```


To quantify these linear associations, Table \@ref(tab:Pairwise-Pearson-correlation-coefficients) represents the pairwise Pearson Correlation Coefficients for the five variables. Attendance has the strongest absolute correlation with Income\_rate ($r=-0.756$), followed by  CIF ($r=-0.735$) and Attainment ($r=0.686$). The association with Crime\_rate is weaker ($r=-0.469$). It is also notable that CIF and Income\_rate have very strongly positive correlation ($r=0.953$), which means substantial overlap. Since Income\_rate has the largest absolute correlation with Attendance among the four selected explanatory variables, it was chosen as the predictor in the simple linear regression model.


### 2.3 Simple Linear Regression

```{r linear-regression-model, echo = FALSE}
# Identify the variable with the highest absolute correlation with Attendance
correlations <- c(
  Attainment = r_attainment,
  Crime_rate = r_crime,
  CIF = r_cif,
  Income_rate = r_income
)
data2_name <- names(which.max(abs(correlations)))

# Fit simple linear regression
model_formula <- as.formula(paste("Attendance ~", data2_name))
lm_model <- lm(model_formula, data = dat)

# Extract coefficients
intercept <- coef(lm_model)[1]
slope     <- coef(lm_model)[2]
r_squared <- summary(lm_model)$r.squared
```

The variable with the highest absolute correlation with Attendance is \textbf{Income Rate} (r=-0.756). We therefore fit a simple linear regression model:

$$\text{Attendance} = \beta_0 + \beta_1\times \text{Income Rate} + \varepsilon$$
The estimated regression equation is: 

$$ \hat{y} = 0.8887 + (-0.635) \times x $$
```{r regression-coef-table, echo = FALSE}
display_name <- gsub("_", " ", data2_name)

knitr::kable(
  tidy(lm_model),
  digits = 4,
  caption = paste("Simple linear regression model: Attendance ~", display_name)
)
```


The estimated coefficients are reported in Table \@ref(tab:regression-coef-table), and their $95\%$ confidence intervals are shown in Table \@ref(tab:regression-ci-table). According the Table \@ref(tab:regression-coef-table), both coefficients are highly statistically significant ($p<0.001$). The estimated slope of $-0.6350$ indicates that higher income deprivation is associated with lower attendance. In particular, a $0.1$ increase in Income\_rate is associated with an average decrease of about $0.0635$ in Attendance.

```{r regression-ci-table, echo = FALSE}
ci <- confint(lm_model, level = 0.95)
rownames(ci) <- c("Intercept ($\\beta_0$)", "Slope ($\\beta_1$)")
colnames(ci) <- c("2.5\\%", "97.5\\%")

knitr::kable(
  ci,
  digits = 4,
  caption = "95\\% Confidence Intervals for Regression Coefficients"
)
```

The $95\%$ confidence interval for the slope is entirely negative, which is consistent with a clear negative linear association between Income_rate and Attendance. The model has \textbf{$R^2=0.5796$}, meaning that about $57.96\%$ of the sample variation in Attendance is accounted by the fitted linear relationship with Income\_rate. This confirms that Income\_rate is an important predictor of Attendance in a simple linear model.


### 2.4 Visualisation of the Fitted Model

```{r linear-regression, echo = FALSE, fig.height=4, fig.width=6, fig.cap='Scatterplot with fitted regression line'}

equation <- paste0("y = ", round(intercept, 3), " + (", round(slope, 3), ") * x")

# Dynamic annotation position
x_pos <- quantile(dat[[data2_name]], 0.75, na.rm = TRUE)
y_pos <- quantile(dat$Attendance, 0.95, na.rm = TRUE)

ggplot(dat, aes(x = !!sym(data2_name), y = Attendance)) +
  geom_point(shape = 16, size = 1, colour = adjustcolor(col_point, alpha.f = 0.5)) +
  geom_smooth(method = "lm", se = TRUE, colour = col_line, fill = adjustcolor(col_line, alpha.f = 0.15)) +
  labs(
    x = data2_name,
    y = "Attendance",
    title = paste("Regression of Attendance on", data2_name)
  ) +
  annotate("text", x = x_pos, y = y_pos, label = equation, parse = FALSE,
           size = 4, colour = col_line) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```


The figure \@ref(fig:linear-regression) shows the scatterplot of Attendance against Income\_rate, together with the fitted regression line and $95\%$ confidence band. The fitted line captures the overall negative linear trend, higher Income\_rate deprivation tend to have lower Attendance. However, there remains noticeable vertical scatter around the line, particular at higher values of Income\_rate. This suggests that additional factors may also influence Attendance.

### 2.5 Residual Diagnostics

```{r residuals, echo = FALSE, fig.height=4, fig.width=6, fig.cap='Residuals of the fitted model'}

# residuals
dat$residuals <- residuals(lm_model)
dat$fitted    <- fitted(lm_model)

ggplot(dat, aes(x = fitted, y = residuals)) +
  geom_point(shape = 16, size = 1, colour = adjustcolor(col_point, alpha.f = 0.5)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = col_ref, linewidth = 0.6) +
  labs(
    x = "Predicted Attendance Rate (%)",
    y = "Residuals",
    title = "Residuals vs Fitted Values"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```


The figure \@ref(fig:residuals) shows the residuals plotted against the fitted values. The residuals are centered around $0$ with no strong systematic pattern and funnel shape, so the linear form appears reasonable. There are a few large negative residuals (data zones with unusually low attendance given their income deprivation level), which could be investigated further.

Overall, among the four variables examined in Section 2, Income\_rate shows the strongest linear association with Attendance and provides the best single predictor linear regression model.
This complements the city level differences found in Section 1 and suggests that socio-economic deprivation may be a key driver of the lower attendance rates observed in Glasgow.

\newpage
### Section 3: [A brief investigation of the student's choosing*]
[Change the title to the investigation you choosing]

[To go here]

\newpage


## Conclusions

We have investigated various characteristics of data available for geographical areas in Scotland called "data zones", focusing in particular ...

[very brief recap of findings]

There are some limitations of the current study. One is ... 

A future study could investigate ...

\newpage


## Appendix

The R code to produce the analysis and plots in this report is as follows:

```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```