---
title: "An Analysis of School Attendance Rates across Scottish Data Zones"
author: "Group XX"
date: ''
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
  bookdown::html_document2:
    number_sections: false
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
dat <- read.csv("scottishData.csv")
library(ggplot2)
library(dplyr)
library(broom)

# Subsets for Section 1
glasgow   <- subset(dat, Council_area == "Glasgow City")
edinburgh <- subset(dat, Council_area == "City of Edinburgh")

# Region variable for Section 3
dat$Region <- ifelse(dat$Council_area == "Glasgow City", "Glasgow",
              ifelse(dat$Council_area == "City of Edinburgh", "Edinburgh", "Elsewhere"))
dat$Region <- factor(dat$Region, levels = c("Glasgow", "Edinburgh", "Elsewhere"))

# ── Unified colour palette ──
col_gla   <- "#D6604D"   # terracotta  — Glasgow
col_edi   <- "#4393C3"   # steel blue  — Edinburgh
col_else  <- "#66C2A5"   # mint green  — Elsewhere
col_point <- "#525252"   # slate gray  — scatter points
col_line  <- "#2166AC"   # deep blue   — regression line
col_ref   <- "#878787"   # medium gray — reference line
```

<!-- ═══════════════════════════════════════════════════════════════
     NOTE FOR WRITING TEAMMATES
     ═══════════════════════════════════════════════════════════════
     Below is the complete Results section (Sections 1–3) with all
     figures, tables, and commentary. You can use this as the basis
     for writing the remaining report sections. Here is a suggested
     outline with key points to cover:

     ## Summary (3 marks)
     - Aim: investigate school attendance across Scottish data zones,
       comparing Glasgow vs Edinburgh, and identifying key predictors.
     - Methods: boxplots, histograms, confidence intervals, t-test,
       F-test (Section 1); correlation, simple linear regression,
       residual diagnostics (Section 2); ANOVA, regional regression,
       interaction model (Section 3).
     - Key findings: Glasgow has significantly lower and more variable
       attendance; Income_rate is the strongest predictor (~58% variance
       explained); the Income_rate effect is weaker in Glasgow than in
       Edinburgh or Elsewhere (interaction significant).

     ## Introduction (5 marks)
     - Context: Scottish data zones, school attendance as indicator of
       well-being and educational engagement.
     - Aim: compare Attendance between Glasgow and Edinburgh, identify
       predictors, and test whether relationships vary by region.
     - Background: Glasgow is known for higher deprivation levels.
     - Data: 400 data zones (100 Glasgow, 100 Edinburgh, 200 Elsewhere),
       34 variables covering demographics, deprivation, health, education.
     - Why it matters: understanding drivers of attendance can inform
       targeted policy interventions.

     ## Methods (3 marks)
     - Section 1: Welch's t-test (unequal variances), F-test for
       variance equality, t-based CI for means, chi-squared CI for
       variances.
     - Section 2: Pearson correlation, simple linear regression (OLS),
       residual analysis for model diagnostics.
     - Section 3: One-way ANOVA, separate regional regressions,
       interaction model (Income_rate × Region) with nested F-test.

     ## Conclusions (4 marks)
     - Recap: Glasgow attendance significantly lower and more variable;
       Income_rate is the dominant predictor; its effect varies by region.
     - Limitations: cross-sectional data (no causation), only one
       predictor in the regression, potential confounders not controlled.
     - Extensions: multiple regression with additional predictors,
       time-series analysis, non-linear models.

     ## Appendix
     - Include all R code (set echo=TRUE or copy-paste from .Rmd files).
     ═══════════════════════════════════════════════════════════════ -->


# Summary Statistics and Hypothesis Test

In this section, we compare the distribution of Attendance rates between Glasgow ($n = 100$) and Edinburgh ($n = 100$), the two largest cities in Scotland, using graphical summaries, descriptive statistics, and formal hypothesis tests.

## Exploratory Visualisation

```{r boxplot, fig.height=4, fig.width=6, fig.cap='Side-by-side boxplots of Attendance for Glasgow and Edinburgh'}

dat_two <- subset(dat, Council_area %in% c("Glasgow City", "City of Edinburgh"))

ggplot(dat_two, aes(x = Council_area, y = Attendance, fill = Council_area)) +
  geom_boxplot(outlier.shape = 1, outlier.size = 1.5) +
  scale_fill_manual(values = c("City of Edinburgh" = col_edi, "Glasgow City" = col_gla)) +
  labs(x = "City", y = "Attendance", title = "Attendance by City") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
```

Figure 1 shows that Edinburgh has a noticeably higher median Attendance than Glasgow. Glasgow also exhibits greater spread, with several low-lying outliers, suggesting that some Glasgow data zones have particularly poor attendance.

```{r histogram, fig.height=4, fig.width=6, fig.cap='Histograms of Attendance for Glasgow and Edinburgh'}

ggplot(dat_two, aes(x = Attendance, fill = Council_area)) +
  geom_histogram(bins = 20, alpha = 0.7, color = "white") +
  scale_fill_manual(values = c("City of Edinburgh" = col_edi, "Glasgow City" = col_gla)) +
  facet_wrap(~ Council_area, ncol = 2) +
  labs(x = "Attendance", y = "Frequency", title = "Distribution of Attendance by City") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
```

Figure 2 reinforces this picture: Edinburgh's distribution is tightly concentrated at higher values, whereas Glasgow's histogram is wider and displays a noticeable left skew. These visual patterns motivate a formal comparison using summary statistics and hypothesis tests below.

## Descriptive Statistics

```{r mean_var}

n_gla <- sum(!is.na(glasgow$Attendance))
n_edi <- sum(!is.na(edinburgh$Attendance))
mean_gla <- mean(glasgow$Attendance, na.rm = TRUE)
mean_edi <- mean(edinburgh$Attendance, na.rm = TRUE)
var_gla  <- var(glasgow$Attendance, na.rm = TRUE)
var_edi  <- var(edinburgh$Attendance, na.rm = TRUE)

sd_gla   <- sd(glasgow$Attendance, na.rm = TRUE)
sd_edi   <- sd(edinburgh$Attendance, na.rm = TRUE)

stats_table <- data.frame(
  City     = c("Glasgow", "Edinburgh"),
  N        = c(n_gla, n_edi),
  Mean     = round(c(mean_gla, mean_edi), 5),
  SD       = round(c(sd_gla, sd_edi), 5),
  Variance = round(c(var_gla, var_edi), 7)
)

knitr::kable(stats_table,
             caption = "Sample means and variances of Attendance for Glasgow and Edinburgh.",
             align = "c")
```

Table 1 confirms the visual impression: Glasgow's mean Attendance (`r round(mean_gla, 4)`) is substantially lower than Edinburgh's (`r round(mean_edi, 4)`), and Glasgow's variance is roughly `r round(var_gla / var_edi, 1)` times larger. These differences suggest that not only does Glasgow have lower average attendance, but its data zones are also far more heterogeneous.

## Confidence Intervals

```{r ci_mean}

ci_mean_gla <- mean_gla + c(-1, 1) * qt(0.975, df = n_gla - 1) * sqrt(var_gla / n_gla)
ci_mean_edi <- mean_edi + c(-1, 1) * qt(0.975, df = n_edi - 1) * sqrt(var_edi / n_edi)

ci_mean_table <- data.frame(
  City  = c("Glasgow", "Edinburgh"),
  Lower = round(c(ci_mean_gla[1], ci_mean_edi[1]), 5),
  Upper = round(c(ci_mean_gla[2], ci_mean_edi[2]), 5)
)

knitr::kable(ci_mean_table,
             caption = "95\\% confidence intervals for the mean of Attendance.",
             align = "c",
             col.names = c("City", "2.5%", "97.5%"))
```

The 95% confidence intervals for the mean (Table 2) do not overlap, providing initial evidence of a significant difference. These intervals are based on the $t$-distribution; given the large sample sizes ($n = 100$), the Central Limit Theorem supports the normality assumption.

```{r ci_var}

ci_var_gla <- c((n_gla - 1) * var_gla / qchisq(0.975, df = n_gla - 1),
                (n_gla - 1) * var_gla / qchisq(0.025, df = n_gla - 1))
ci_var_edi <- c((n_edi - 1) * var_edi / qchisq(0.975, df = n_edi - 1),
                (n_edi - 1) * var_edi / qchisq(0.025, df = n_edi - 1))

ci_var_table <- data.frame(
  City  = c("Glasgow", "Edinburgh"),
  Lower = round(c(ci_var_gla[1], ci_var_edi[1]), 7),
  Upper = round(c(ci_var_gla[2], ci_var_edi[2]), 7)
)

knitr::kable(ci_var_table,
             caption = "95\\% confidence intervals for the variance of Attendance.",
             align = "c",
             col.names = c("City", "2.5%", "97.5%"))
```

The 95% confidence intervals for the variance (Table 3) show that Glasgow's interval lies entirely above Edinburgh's, confirming that the dispersion in Glasgow is significantly greater. These intervals rely on the $\chi^2$ distribution and hence assume approximate normality.

Taken together, the non-overlapping confidence intervals for both the mean and the variance strongly suggest real differences between the two cities. We now verify this with formal hypothesis tests.

## Hypothesis Tests

**Test for difference in means.**

```{r t_test}

t_result <- t.test(glasgow$Attendance, edinburgh$Attendance, var.equal = FALSE)

t_table <- data.frame(
  Statistic = round(t_result$statistic, 4),
  df        = round(t_result$parameter, 2),
  p_value   = formatC(t_result$p.value, format = "e", digits = 4)
)

knitr::kable(t_table,
             caption = "Welch's two-sample t-test for difference in mean Attendance.",
             align = "c",
             col.names = c("t-statistic", "df", "p-value"))
```

We test $H_0: \mu_{Glasgow} = \mu_{Edinburgh}$ against $H_1: \mu_{Glasgow} \neq \mu_{Edinburgh}$ using Welch's two-sample $t$-test, which does not require equal variances. The test yields $t =$ `r round(t_result$statistic, 3)` with $p$-value $=$ `r formatC(t_result$p.value, format = "e", digits = 3)`. Since $p < 0.05$, we reject $H_0$ and conclude that Glasgow's mean Attendance is significantly lower than Edinburgh's.

**Test for difference in variability.**

```{r f_test}

f_result <- var.test(glasgow$Attendance, edinburgh$Attendance)

f_table <- data.frame(
  Statistic = round(f_result$statistic, 4),
  df1       = f_result$parameter[1],
  df2       = f_result$parameter[2],
  p_value   = formatC(f_result$p.value, format = "e", digits = 4)
)

knitr::kable(f_table,
             caption = "F-test for equality of variances of Attendance.",
             align = "c",
             col.names = c("F-statistic", "df1", "df2", "p-value"))
```

We test $H_0: \sigma^2_{Glasgow} = \sigma^2_{Edinburgh}$ against $H_1: \sigma^2_{Glasgow} \neq \sigma^2_{Edinburgh}$ using an $F$-test (which assumes normality). The result is $F =$ `r round(f_result$statistic, 3)` with $p$-value $=$ `r formatC(f_result$p.value, format = "e", digits = 3)`. Since $p < 0.05$, we reject $H_0$ and conclude that Glasgow's Attendance is significantly more variable than Edinburgh's.

In summary, both the central tendency and the spread of Attendance differ significantly between Glasgow and Edinburgh. Glasgow has both lower and more variable attendance rates, which may reflect underlying socio-economic disparities. These findings motivate a deeper investigation in Section 2, where we explore which factors are most strongly associated with Attendance across all Scottish data zones.


# Correlation and Linear Regression

Having established that Attendance differs between Glasgow and Edinburgh, we now consider all 400 data zones together to investigate which socio-economic factors are most closely associated with Attendance. We select four candidate predictors: Attainment, crime_rate, CIF (Community Integration Factor), and Income_rate.

## Pairwise Relationships

```{r scatters_1, fig.height=6, fig.width=7, fig.cap='Pairwise scatterplot matrix of Attendance and four selected variables'}

vars <- c("Attendance", "Attainment", "crime_rate", "CIF", "Income_rate")

pairs(dat[, vars],
      pch = 16, cex = 0.4, col = adjustcolor(col_point, alpha.f = 0.5),
      main = "",
      labels = c("Attendance", "Attainment", "Crime Rate", "CIF", "Income Rate"))
```

Figure 3 provides an overview of the bivariate relationships. Attendance appears negatively associated with Income_rate, CIF, and crime_rate, while displaying a positive trend with Attainment. The strongest visual pattern is the tight negative relationship between Attendance and Income_rate. To quantify these associations, we compute the full matrix of Pearson correlation coefficients.

## Pearson Correlation Coefficients

```{r Pairwise_Pearson_correlation_coefficients}
# Full 5x5 pairwise Pearson correlation matrix
cor_matrix <- round(cor(dat[, vars], use = "complete.obs", method = "pearson"), 3)
colnames(cor_matrix) <- c("Attendance", "Attainment", "Crime Rate", "CIF", "Income Rate")
rownames(cor_matrix) <- c("Attendance", "Attainment", "Crime Rate", "CIF", "Income Rate")

knitr::kable(cor_matrix,
             caption = "Pairwise Pearson correlation coefficients between the five selected variables.",
             align = "c")

r_attainment <- cor_matrix[1, 2]
r_crime      <- cor_matrix[1, 3]
r_cif        <- cor_matrix[1, 4]
r_income     <- cor_matrix[1, 5]
```

Table 6 confirms the visual impression. Attendance is most strongly correlated with Income_rate ($r =$ `r r_income`), followed by CIF ($r =$ `r r_cif`) and Attainment ($r =$ `r r_attainment`). Crime_rate shows a weaker association ($r =$ `r r_crime`). Notably, Income_rate, CIF, and crime_rate are also highly inter-correlated, reflecting the well-known overlap between deprivation indicators.

Since Income_rate has the highest absolute correlation with Attendance, we select it as the predictor for a simple linear regression model.

## Simple Linear Regression

```{r Linear_Regression}
correlations <- c(
  Attainment = r_attainment,
  Crime_rate = r_crime,
  CIF = r_cif,
  Income_rate = r_income
)
data2_name <- names(which.max(abs(correlations)))

model_formula <- as.formula(paste("Attendance ~", data2_name))
lm_model <- lm(model_formula, data = dat)

intercept <- coef(lm_model)[1]
slope     <- coef(lm_model)[2]
r_squared <- summary(lm_model)$r.squared
```

The variable with the highest absolute correlation with Attendance is **`r data2_name`** ($r =$ `r round(correlations[data2_name], 3)`). We therefore fit a simple linear regression model:
$$\text{Attendance} = \beta_0 + \beta_1 \times \text{`r gsub("_", "\\\\_", data2_name)`} + \varepsilon$$

The estimated regression equation is: $\hat{y} =$ `r round(intercept, 4)` $+$ (`r round(slope, 4)`) $\times x$.

```{r Linear_Regression_Analysis}

knitr::kable(tidy(lm_model), digits = 4,
             caption = paste("Simple linear regression model: Attendance ~", data2_name))

ci <- confint(lm_model, level = 0.95)
rownames(ci) <- c("Intercept ($\\beta_0$)", "Slope ($\\beta_1$)")
colnames(ci) <- c("2.5%", "97.5%")
knitr::kable(ci, digits = 4, caption = "95\\% Confidence Intervals for Regression Coefficients")
```

Both coefficients are highly significant ($p < 0.001$). The slope $\hat{\beta}_1 =$ `r round(slope, 4)` indicates that, on average, a one-unit increase in `r data2_name` is associated with a `r round(abs(slope), 4)` decrease in Attendance. The model explains `r round(r_squared * 100, 2)`% of the total variance ($R^2 =$ `r round(r_squared, 4)`), suggesting that `r data2_name` alone is a strong predictor of Attendance.

## Visualisation of the Fitted Model

```{r linear_regression_plot, fig.height=4, fig.width=6, fig.cap='Scatterplot with fitted regression line'}

equation <- paste0("y = ", round(intercept, 3), " + (", round(slope, 3), ") * x")

x_pos <- quantile(dat[[data2_name]], 0.75, na.rm = TRUE)
y_pos <- quantile(dat$Attendance, 0.95, na.rm = TRUE)

ggplot(dat, aes(x = !!sym(data2_name), y = Attendance)) +
  geom_point(shape = 16, size = 1, colour = adjustcolor(col_point, alpha.f = 0.5)) +
  geom_smooth(method = "lm", se = TRUE, colour = col_line, fill = adjustcolor(col_line, alpha.f = 0.15)) +
  labs(x = data2_name, y = "Attendance",
       title = paste("Regression of Attendance on", data2_name)) +
  annotate("text", x = x_pos, y = y_pos, label = equation, parse = FALSE,
           size = 4, colour = col_line) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

Figure 4 visualises the fitted line superimposed on the raw data. The negative trend is clear, and the shaded 95% confidence band for the regression line is narrow, reflecting the large sample size. Some dispersion is evident at higher values of `r data2_name`, which we examine further via residual diagnostics.

## Residual Diagnostics

```{r residuals_plot, fig.height=4, fig.width=6, fig.cap='Residuals of the fitted model'}

dat$residuals <- residuals(lm_model)
dat$fitted    <- fitted(lm_model)

ggplot(dat, aes(x = fitted, y = residuals)) +
  geom_point(shape = 16, size = 1, colour = adjustcolor(col_point, alpha.f = 0.5)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = col_ref, linewidth = 0.6) +
  labs(x = "Predicted Attendance Rate (%)", y = "Residuals",
       title = "Residuals vs Fitted Values") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

The residual plot (Figure 5) shows residuals scattered around zero with no obvious systematic pattern or funnel shape, indicating that the linearity and constant-variance assumptions are reasonably satisfied. There are a few large negative residuals (data zones with unusually low attendance given their income deprivation level), which could be investigated further.

Overall, the results of Section 2 demonstrate that income deprivation is the single strongest linear predictor of attendance among the variables considered, explaining approximately `r round(r_squared * 100, 0)`% of the variation. This complements the city-level differences found in Section 1 and suggests that socio-economic deprivation may be a key driver of the lower attendance rates observed in Glasgow. An important follow-up question is whether this relationship holds equally across regions — we investigate this in Section 3.


# A Further Investigation

Sections 1 and 2 revealed that (i) Glasgow has significantly lower and more variable Attendance than Edinburgh, and (ii) Income_rate is the strongest single predictor of Attendance across all data zones. A natural question follows: **does the relationship between income deprivation and attendance differ across Glasgow, Edinburgh, and the rest of Scotland?** We investigate this by comparing distributions, fitting regional regression models, and formally testing for interaction effects.

## Regional Overview

We first compare Attendance across the three regions visually and then test for differences using a one-way ANOVA.

```{r boxplot_region, fig.height=4, fig.width=6, fig.cap='Distribution of Attendance across Glasgow, Edinburgh, and Elsewhere'}

ggplot(dat, aes(x = Region, y = Attendance, fill = Region)) +
  geom_boxplot(outlier.shape = 1, outlier.size = 1.5) +
  scale_fill_manual(values = c("Glasgow" = col_gla, "Edinburgh" = col_edi, "Elsewhere" = col_else)) +
  labs(x = "Region", y = "Attendance", title = "Attendance by Region") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
```

```{r anova_region}

region_stats <- dat %>%
  group_by(Region) %>%
  summarise(N = n(),
            Mean = round(mean(Attendance, na.rm = TRUE), 4),
            SD   = round(sd(Attendance, na.rm = TRUE), 4),
            .groups = "drop")

knitr::kable(region_stats,
             caption = "Descriptive statistics of Attendance by region.",
             align = "c")

aov_result <- aov(Attendance ~ Region, data = dat)
aov_summary <- summary(aov_result)
f_stat <- round(aov_summary[[1]]$`F value`[1], 2)
p_aov  <- aov_summary[[1]]$`Pr(>F)`[1]
```

Figure 6 shows that Glasgow has the lowest median Attendance and the widest spread, while Edinburgh and Elsewhere have similar, higher medians. A one-way ANOVA yields $F =$ `r f_stat` with $p$-value $=$ `r formatC(p_aov, format = "e", digits = 3)`, confirming that mean Attendance differs significantly across the three regions.

## Regional Correlation and Regression

We now examine whether the Income_rate–Attendance relationship identified in Section 2 holds equally across all three regions.

```{r correlation_region}

cor_by_region <- sapply(split(dat, dat$Region), function(x)
  round(cor(x$Attendance, x$Income_rate, method = "pearson"), 3))

cor_results <- data.frame(
  Region = names(cor_by_region),
  Pearson_r = as.numeric(cor_by_region)
)

knitr::kable(cor_results,
             caption = "Pearson correlation between Income\\_rate and Attendance by region.",
             align = "c")
```

The correlations are negative in all three regions, confirming that income deprivation is universally associated with lower attendance. However, the strength varies: Glasgow ($r =$ `r cor_by_region["Glasgow"]`) shows a noticeably weaker association than Edinburgh ($r =$ `r cor_by_region["Edinburgh"]`) and Elsewhere ($r =$ `r cor_by_region["Elsewhere"]`).

```{r linear_regression_region}

region_data <- split(dat, dat$Region)
models <- lapply(region_data, function(x) lm(Attendance ~ Income_rate, data = x))

intercepts <- sapply(models, function(m) coef(m)[1])
slopes     <- sapply(models, function(m) coef(m)[2])
r_sq       <- sapply(models, function(m) summary(m)$r.squared)

reg_results <- data.frame(
  Region    = names(models),
  Intercept = round(intercepts, 3),
  Slope     = round(slopes, 3),
  R_squared = round(r_sq, 3)
)

knitr::kable(reg_results,
             caption = "Regional regression models: Attendance $\\sim$ Income\\_rate.",
             align = "c",
             col.names = c("Region", "$\\hat{\\beta}_0$", "$\\hat{\\beta}_1$", "$R^2$"))
```

In Glasgow, income deprivation explains only `r round(r_sq["Glasgow"] * 100, 1)`% of the variance ($R^2 =$ `r round(r_sq["Glasgow"], 3)`), with a relatively shallow slope of `r round(slopes["Glasgow"], 3)`. By contrast, the slopes are steeper in Edinburgh (`r round(slopes["Edinburgh"], 3)`) and Elsewhere (`r round(slopes["Elsewhere"], 3)`), with $R^2$ values of `r round(r_sq["Edinburgh"], 3)` and `r round(r_sq["Elsewhere"], 3)` respectively. The scatterplot below illustrates these differences.

```{r scatter_region, fig.height=4, fig.width=6, fig.cap='Regional regression lines for Attendance on Income\\_rate'}

ggplot(dat, aes(x = Income_rate, y = Attendance, colour = Region)) +
  geom_point(shape = 16, size = 1, alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8, alpha = 0.15) +
  scale_colour_manual(values = c("Glasgow" = col_gla, "Edinburgh" = col_edi, "Elsewhere" = col_else)) +
  labs(x = "Income deprivation rate", y = "Attendance",
       title = "Attendance vs Income Rate by Region") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

The three regression lines have noticeably different slopes: Glasgow's line is flatter, whereas Edinburgh and Elsewhere decline more steeply.

## Testing for Regional Differences in Slope

To formally test whether the effect of Income_rate on Attendance differs by region, we fit an interaction model:

$$\text{Attendance} = \beta_0 + \beta_1 \times \text{Income\_rate} + \beta_2 \times \text{Region} + \beta_3 \times (\text{Income\_rate} \times \text{Region}) + \varepsilon$$

```{r interaction_model}

int_model <- lm(Attendance ~ Income_rate * Region, data = dat)

knitr::kable(tidy(int_model), digits = 4,
             caption = "Interaction model: Attendance $\\sim$ Income\\_rate $\\times$ Region.")

add_model <- lm(Attendance ~ Income_rate + Region, data = dat)
anova_int <- anova(add_model, int_model)
f_int <- round(anova_int$F[2], 3)
p_int <- anova_int$`Pr(>F)`[2]
r2_int <- round(summary(int_model)$r.squared, 4)
```

The interaction terms test whether the slope of Income_rate differs significantly between regions. An $F$-test comparing the additive model (no interaction) against the interaction model yields $F =$ `r f_int` with $p =$ `r formatC(p_int, format = "e", digits = 3)`. Since $p < 0.05$, the interaction is statistically significant: **the effect of income deprivation on attendance is not uniform across regions.** The full interaction model explains `r round(r2_int * 100, 2)`% of the variance in Attendance.

In particular, the significantly positive interaction coefficient for Glasgow indicates that Glasgow's slope is less steep (less negative) than the baseline category, consistent with the weaker relationship observed in the regional regressions above.

## Summary

The analysis reveals three key findings:

1. Attendance differs significantly across Glasgow, Edinburgh, and Elsewhere (ANOVA $p <$ 0.001).
2. Income deprivation is negatively associated with attendance in all three regions, but the relationship is substantially weaker in Glasgow ($R^2 \approx$ `r round(r_sq["Glasgow"], 2)`) than in Edinburgh ($R^2 \approx$ `r round(r_sq["Edinburgh"], 2)`) or Elsewhere ($R^2 \approx$ `r round(r_sq["Elsewhere"], 2)`).
3. A formal interaction test confirms that the slope of Income_rate on Attendance differs significantly by region ($p =$ `r formatC(p_int, format = "e", digits = 3)`).

These results suggest that while income deprivation is an important universal predictor of attendance, **its explanatory power varies geographically**. In Glasgow, other unmeasured factors — such as local education policies, community-level interventions, or housing conditions — may play a comparatively larger role in shaping attendance outcomes. Future work could explore multiple regression with additional predictors to explain the residual variation in Glasgow.
