---
title: "An Analysis of School Attendance Rates across Scottish Data Zones"
output:
  pdf_document:
    latex_engine: xelatex
---

Section 3 extends the findings of Section 2 by examining whether the relationship between income deprivation and school attendance differs across Glasgow, Edinburgh, and the rest of Scotland. 

```{r correlation, echo=FALSE, warning=FALSE}
dat <- read.csv("scottishData.csv")

# Create region variable
dat$Region <- ifelse(dat$Council_area == "Glasgow City", "Glasgow",
              ifelse(dat$Council_area == "City of Edinburgh", "Edinburgh", "Elsewhere"))
dat$Region <- factor(dat$Region, levels = c("Glasgow", "Edinburgh", "Elsewhere"))

# Pearson correlations by region
cor_results <- data.frame(
  Region = levels(dat$Region),
  Pearson_Correlation = sapply(split(dat, dat$Region), function(x)
    cor(x$Attendance, x$Income_rate, method = "pearson"))
)

cor_results$Pearson_Correlation <- round(cor_results$Pearson_Correlation, 3)

knitr::kable(cor_results,
             caption = "Pearson correlation between Income_rate and Attendance by region",
             align = "c"
             )
```

The Pearson correlation coefficients are negative in all three regions, confirming that higher levels of income deprivation are consistently associated with lower attendance rates. However, the strength of this relationship varies substantially. Glasgow shows the weakest correlation ($r = -0.628$), while Edinburgh ($r = -0.829$) and Elsewhere Scotland ($r = -0.811$) display much stronger negative associations. This suggests that income deprivation is a particularly strong indicator of attendance outside Glasgow.

```{r linear_regression, echo=FALSE, warning=FALSE}

region_data <- split(dat, dat$Region)
models <- lapply(region_data, function(x)
  lm(Attendance ~ Income_rate, data = x))
intercept <- sapply(models, function(m) coef(m)[1])
slope     <- sapply(models, function(m) coef(m)[2])

r_squared <- sapply(models, function(m) summary(m)$r.squared)
reg_results <- data.frame(
  Region = names(models),
  Equation = paste0(
    "Attendance = ",
    round(intercept, 3),
    ifelse(slope >= 0, " + ", " - "),
    round(abs(slope), 3),
    " × Income_rate"
  ),
  
  R_squared = round(r_squared, 3)
)

knitr::kable(
  reg_results,
  caption = "Regional regression equations for Attendance on Income_rate",
  align = "c"
)
```

The regional regression models reinforce this conclusion. In Glasgow, the fitted equation is $Attendance = 0.864 − 0.455 \times Income_rate$, with $R^{2} = 0.395$, meaning income deprivation explains only around $40\%$ of the variation in attendance. By contrast, the slopes are much steeper in Edinburgh ($-0.822$) and Elsewhere ($-0.778$), with substantially higher explanatory power ($R^{2} = 0.688$ and $0.657$, respectively). The scatterplot below visually supports these results, as the fitted lines for Edinburgh and Elsewhere decline more sharply than that of Glasgow.

```{r scatter, echo=FALSE, warning=FALSE, fig.height=4, fig.width=6, fig.cap='Relationship between Income_rate and Attendance by Region'}
library(ggplot2)
ggplot(dat, aes(x = Income_rate, y = Attendance, colour = Region)) +
  geom_point(shape = 1, size = 1, stroke = 0.5, alpha = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
    
  scale_colour_manual(values = c(
    "Glasgow" = "steelblue4",
    "Edinburgh" = "dodgerblue3",
    "Elsewhere" = "skyblue3"
    ))+

  labs(
    title = "Relationship between Income_rate and Attendance by Region",
    x = "Income deprivation rate",
    y = "Attendance"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank(),
    align = "c"
  )

```

Overall, while income deprivation negatively affects attendance in all regions, it appears to be a weaker predictor in Glasgow. This may indicate that additional social, educational, or demographic factors play a more important role in explaining attendance differences within Glasgow.
