---
title: "An Analysis of School Attendance Rates across Scottish Data Zones"
date: ''
output:
  pdf_document: default
  bookdown::html_document2:
    number_sections: false
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
dat <- read.csv("../scottishData.csv")

glasgow   <- subset(dat, Council_area == "Glasgow City")
edinburgh <- subset(dat, Council_area == "City of Edinburgh")

# Unified colour palette
col_gla <- "#D6604D"
col_edi <- "#4393C3"
```


# Section 1: Summary Statistics and Hypothesis Test

In this section, we compare the distribution of Attendance rates between Glasgow ($n = 100$) and Edinburgh ($n = 100$), the two largest cities in Scotland, using graphical summaries, descriptive statistics, and formal hypothesis tests.

## 1.1 Exploratory Visualisation

```{r boxplot, echo=FALSE, fig.height=4, fig.width=6, fig.cap='Side-by-side boxplots of Attendance for Glasgow and Edinburgh'}

boxplot(glasgow$Attendance, edinburgh$Attendance,
        names = c("Glasgow", "Edinburgh"),
        col = c(adjustcolor(col_gla, 0.6), adjustcolor(col_edi, 0.6)),
        border = c(col_gla, col_edi),
        ylab = "Attendance (proportion)",
        main = "Attendance by City",
        pch = 1, cex = 0.8)
```

Figure 1 shows that Edinburgh has a noticeably higher median Attendance than Glasgow. Glasgow also exhibits greater spread, with several low-lying outliers, suggesting that some Glasgow data zones have particularly poor attendance.

```{r histogram, echo=FALSE, fig.height=4, fig.width=6, fig.cap='Histograms of Attendance for Glasgow and Edinburgh'}

par(mfrow = c(1, 2))
hist(glasgow$Attendance, breaks = 20, col = adjustcolor(col_gla, 0.6),
     border = "white", main = "Glasgow", xlab = "Attendance (proportion)", ylab = "Frequency")
abline(v = mean(glasgow$Attendance), col = col_gla, lwd = 2, lty = 2)
hist(edinburgh$Attendance, breaks = 20, col = adjustcolor(col_edi, 0.6),
     border = "white", main = "Edinburgh", xlab = "Attendance (proportion)", ylab = "Frequency")
abline(v = mean(edinburgh$Attendance), col = col_edi, lwd = 2, lty = 2)
legend("topleft", legend = "Sample mean", lty = 2, lwd = 2, col = "black", cex = 0.75, bty = "n")
```

Figure 2 reinforces this picture: Edinburgh's distribution is tightly concentrated at higher values, whereas Glasgow's histogram is wider and displays a noticeable left skew. These visual patterns motivate a formal comparison using summary statistics and hypothesis tests below.

## 1.2 Descriptive Statistics

```{r mean_var, echo = FALSE}

n_gla <- sum(!is.na(glasgow$Attendance))
n_edi <- sum(!is.na(edinburgh$Attendance))
mean_gla <- mean(glasgow$Attendance, na.rm = TRUE)
mean_edi <- mean(edinburgh$Attendance, na.rm = TRUE)
var_gla  <- var(glasgow$Attendance, na.rm = TRUE)
var_edi  <- var(edinburgh$Attendance, na.rm = TRUE)

sd_gla   <- sd(glasgow$Attendance, na.rm = TRUE)
sd_edi   <- sd(edinburgh$Attendance, na.rm = TRUE)

stats_table <- data.frame(
  City     = c("Glasgow", "Edinburgh"),
  N        = c(n_gla, n_edi),
  Mean     = round(c(mean_gla, mean_edi), 5),
  SD       = round(c(sd_gla, sd_edi), 5),
  Variance = round(c(var_gla, var_edi), 7)
)

knitr::kable(stats_table,
             caption = "Sample means and variances of Attendance for Glasgow and Edinburgh.",
             align = "c")
```

Table 1 confirms the visual impression: Glasgow's mean Attendance (`r round(mean_gla, 4)`) is substantially lower than Edinburgh's (`r round(mean_edi, 4)`), and Glasgow's variance is roughly `r round(var_gla / var_edi, 1)` times larger. These differences suggest that not only does Glasgow have lower average attendance, but its data zones are also far more heterogeneous.

## 1.3 Confidence Intervals

```{r ci_mean, echo = FALSE}

# 95% CI for means using t-distribution
ci_mean_gla <- mean_gla + c(-1, 1) * qt(0.975, df = n_gla - 1) * sqrt(var_gla / n_gla)
ci_mean_edi <- mean_edi + c(-1, 1) * qt(0.975, df = n_edi - 1) * sqrt(var_edi / n_edi)

ci_mean_table <- data.frame(
  City  = c("Glasgow", "Edinburgh"),
  Lower = round(c(ci_mean_gla[1], ci_mean_edi[1]), 5),
  Upper = round(c(ci_mean_gla[2], ci_mean_edi[2]), 5)
)

knitr::kable(ci_mean_table,
             caption = "95% confidence intervals for the mean of Attendance.",
             align = "c",
             col.names = c("City", "2.5%", "97.5%"))
```

The 95% confidence intervals for the mean (Table 2) do not overlap, providing initial evidence of a significant difference. These intervals are based on the $t$-distribution; given the large sample sizes ($n = 100$), the Central Limit Theorem supports the normality assumption.

```{r ci_var, echo = FALSE}

# 95% CI for variance using chi-squared distribution
ci_var_gla <- c((n_gla - 1) * var_gla / qchisq(0.975, df = n_gla - 1),
                (n_gla - 1) * var_gla / qchisq(0.025, df = n_gla - 1))
ci_var_edi <- c((n_edi - 1) * var_edi / qchisq(0.975, df = n_edi - 1),
                (n_edi - 1) * var_edi / qchisq(0.025, df = n_edi - 1))

ci_var_table <- data.frame(
  City  = c("Glasgow", "Edinburgh"),
  Lower = round(c(ci_var_gla[1], ci_var_edi[1]), 7),
  Upper = round(c(ci_var_gla[2], ci_var_edi[2]), 7)
)

knitr::kable(ci_var_table,
             caption = "95% confidence intervals for the variance of Attendance.",
             align = "c",
             col.names = c("City", "2.5%", "97.5%"))
```

The 95% confidence intervals for the variance (Table 3) show that Glasgow's interval lies entirely above Edinburgh's, confirming that the dispersion in Glasgow is significantly greater. These intervals rely on the $\chi^2$ distribution and hence assume approximate normality.

Taken together, the non-overlapping confidence intervals for both the mean and the variance strongly suggest real differences between the two cities. We now verify this with formal hypothesis tests.

## 1.4 Hypothesis Tests

**Test for difference in means.**

```{r t_test, echo = FALSE}

# Manual t-statistic calculation (Welch's t-test, as in Workshop 2)
t_stat <- (mean_gla - mean_edi) / sqrt(var_gla / n_gla + var_edi / n_edi)
df_welch <- (var_gla / n_gla + var_edi / n_edi)^2 /
            ((var_gla / n_gla)^2 / (n_gla - 1) + (var_edi / n_edi)^2 / (n_edi - 1))
p_val_t <- 2 * pt(abs(t_stat), df = df_welch, lower.tail = FALSE)

# Verify with built-in function
t_result <- t.test(glasgow$Attendance, edinburgh$Attendance, var.equal = FALSE)

t_table <- data.frame(
  Statistic = round(t_result$statistic, 4),
  df        = round(t_result$parameter, 2),
  p_value   = formatC(t_result$p.value, format = "e", digits = 4)
)

knitr::kable(t_table,
             caption = "Welch's two-sample t-test for difference in mean Attendance.",
             align = "c",
             col.names = c("t-statistic", "df", "p-value"))
```

We test $H_0: \mu_{Glasgow} = \mu_{Edinburgh}$ against $H_1: \mu_{Glasgow} \neq \mu_{Edinburgh}$ using Welch's two-sample $t$-test, which does not require equal variances. The test yields $t =$ `r round(t_result$statistic, 3)` with $p$-value $=$ `r formatC(t_result$p.value, format = "e", digits = 3)`. Since $p < 0.05$, we reject $H_0$ and conclude that Glasgow's mean Attendance is significantly lower than Edinburgh's.

**Test for difference in variability.**

```{r f_test, echo = FALSE}

# Manual F-statistic calculation (as in Workshop 4)
f_stat_manual <- var_gla / var_edi
p_val_f <- 2 * min(pf(f_stat_manual, n_gla - 1, n_edi - 1, lower.tail = FALSE),
                   pf(f_stat_manual, n_gla - 1, n_edi - 1, lower.tail = TRUE))

# Verify with built-in function
f_result <- var.test(glasgow$Attendance, edinburgh$Attendance)

f_table <- data.frame(
  Statistic = round(f_result$statistic, 4),
  df1       = f_result$parameter[1],
  df2       = f_result$parameter[2],
  p_value   = formatC(f_result$p.value, format = "e", digits = 4)
)

knitr::kable(f_table,
             caption = "F-test for equality of variances of Attendance.",
             align = "c",
             col.names = c("F-statistic", "df1", "df2", "p-value"))
```

We test $H_0: \sigma^2_{Glasgow} = \sigma^2_{Edinburgh}$ against $H_1: \sigma^2_{Glasgow} \neq \sigma^2_{Edinburgh}$ using an $F$-test (which assumes normality). The result is $F =$ `r round(f_result$statistic, 3)` with $p$-value $=$ `r formatC(f_result$p.value, format = "e", digits = 3)`. Since $p < 0.05$, we reject $H_0$ and conclude that Glasgow's Attendance is significantly more variable than Edinburgh's.

In summary, both the central tendency and the spread of Attendance differ significantly between Glasgow and Edinburgh. Glasgow has both lower and more variable attendance rates, which may reflect underlying socio-economic disparities. These findings motivate a deeper investigation in Section 2, where we explore which factors are most strongly associated with Attendance across all Scottish data zones.
