---
title: "An Analysis of School Attendance Rates across Scottish Data Zones"
date: ''
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
dat <- read.csv("../scottishData.csv")

# Create region variable
dat$Region <- ifelse(dat$Council_area == "Glasgow City", "Glasgow",
              ifelse(dat$Council_area == "City of Edinburgh", "Edinburgh", "Elsewhere"))
dat$Region <- factor(dat$Region, levels = c("Glasgow", "Edinburgh", "Elsewhere"))

# Subsets
glasgow   <- subset(dat, Region == "Glasgow")
edinburgh <- subset(dat, Region == "Edinburgh")
elsewhere <- subset(dat, Region == "Elsewhere")

# Unified colour palette
col_gla   <- "#D6604D"   # terracotta  — Glasgow
col_edi   <- "#4393C3"   # steel blue  — Edinburgh
col_else  <- "#66C2A5"   # mint green  — Elsewhere
col_point <- "#525252"   # slate gray  — scatter points
col_line  <- "#2166AC"   # deep blue   — regression line
col_ref   <- "#878787"   # medium gray — reference line
```


# Section 3: A Further Investigation

Sections 1 and 2 revealed that (i) Glasgow has significantly lower and more variable Attendance than Edinburgh, and (ii) Income_rate is the strongest single predictor of Attendance across all 400 data zones. A natural follow-up question is: **is Glasgow's lower attendance fully explained by its higher income deprivation, or does a residual "Glasgow effect" persist?** We address this by extending our analysis to three regions — Glasgow, Edinburgh, and the rest of Scotland — and applying a broader range of statistical methods.

## 3.1 Regional Overview

```{r boxplot_3region, echo=FALSE, fig.height=4, fig.width=6, fig.cap='Distribution of Attendance across three regions'}

boxplot(glasgow$Attendance, edinburgh$Attendance, elsewhere$Attendance,
        names = c("Glasgow", "Edinburgh", "Elsewhere"),
        col = c(adjustcolor(col_gla), adjustcolor(col_edi), adjustcolor(col_else)),
        border = c(col_gla, col_edi, col_else),
        ylab = "Attendance (proportion)",
        main = "Attendance by Region",
        pch = 1, cex = 0.8)
```

```{r desc_stats_3region, echo=FALSE}

n_gla  <- length(glasgow$Attendance)
n_edi  <- length(edinburgh$Attendance)
n_else <- length(elsewhere$Attendance)

mean_gla  <- mean(glasgow$Attendance)
mean_edi  <- mean(edinburgh$Attendance)
mean_else <- mean(elsewhere$Attendance)

var_gla  <- var(glasgow$Attendance)
var_edi  <- var(edinburgh$Attendance)
var_else <- var(elsewhere$Attendance)

stats_table <- data.frame(
  Region   = c("Glasgow", "Edinburgh", "Elsewhere"),
  N        = c(n_gla, n_edi, n_else),
  Mean     = round(c(mean_gla, mean_edi, mean_else), 5),
  SD       = round(c(sd(glasgow$Attendance), sd(edinburgh$Attendance), sd(elsewhere$Attendance)), 5),
  Variance = round(c(var_gla, var_edi, var_else), 7)
)

knitr::kable(stats_table,
             caption = "Descriptive statistics of Attendance by region.",
             align = "c")
```

Figure 6 shows that Glasgow has the lowest median Attendance and the widest spread. Edinburgh is tightly concentrated at higher values, while Elsewhere falls between the two. Table 8 confirms this pattern numerically.

## 3.2 Pairwise Hypothesis Tests

We perform pairwise comparisons between Glasgow and Elsewhere (the 200 data zones outside both major cities).

**Test for difference in means.**

```{r t_tests_pairwise, echo=FALSE}

# Glasgow vs Elsewhere
t_gla_else <- t.test(glasgow$Attendance, elsewhere$Attendance, var.equal = FALSE)
# Edinburgh vs Elsewhere
t_edi_else <- t.test(edinburgh$Attendance, elsewhere$Attendance, var.equal = FALSE)

t_table <- data.frame(
  Comparison = c("Glasgow vs Elsewhere", "Edinburgh vs Elsewhere"),
  t_stat     = round(c(t_gla_else$statistic, t_edi_else$statistic), 4),
  df         = round(c(t_gla_else$parameter, t_edi_else$parameter), 2),
  p_value    = c(formatC(t_gla_else$p.value, format = "e", digits = 4),
                 formatC(t_edi_else$p.value, format = "e", digits = 4))
)

knitr::kable(t_table,
             caption = "Welch's t-tests comparing mean Attendance between regions.",
             align = "c",
             col.names = c("Comparison", "t-statistic", "df", "p-value"))
```

Both $p$-values are well below 0.05, indicating that Glasgow's mean Attendance is significantly lower than Elsewhere, and Edinburgh's is significantly higher.

**Test for difference in variability.**

```{r f_tests_pairwise, echo=FALSE}

# Glasgow vs Elsewhere
f_gla_else <- var.test(glasgow$Attendance, elsewhere$Attendance)
# Edinburgh vs Elsewhere
f_edi_else <- var.test(edinburgh$Attendance, elsewhere$Attendance)

f_table <- data.frame(
  Comparison = c("Glasgow vs Elsewhere", "Edinburgh vs Elsewhere"),
  F_stat     = round(c(f_gla_else$statistic, f_edi_else$statistic), 4),
  df1        = c(f_gla_else$parameter[1], f_edi_else$parameter[1]),
  df2        = c(f_gla_else$parameter[2], f_edi_else$parameter[2]),
  p_value    = c(formatC(f_gla_else$p.value, format = "e", digits = 4),
                 formatC(f_edi_else$p.value, format = "e", digits = 4))
)

knitr::kable(f_table,
             caption = "F-tests comparing variance of Attendance between regions.",
             align = "c",
             col.names = c("Comparison", "F-statistic", "df1", "df2", "p-value"))
```

Glasgow is significantly more variable than Elsewhere, while Edinburgh is significantly less variable. This confirms that the patterns observed in Section 1 are not unique to the Glasgow–Edinburgh comparison but extend to the wider Scottish context.

## 3.3 Regional Regression Analysis

We fit separate simple linear regression models (Attendance $\sim$ Income\_rate) for each region and compare slopes.

```{r regional_regression, echo=FALSE}

lm_gla  <- lm(Attendance ~ Income_rate, data = glasgow)
lm_edi  <- lm(Attendance ~ Income_rate, data = edinburgh)
lm_else <- lm(Attendance ~ Income_rate, data = elsewhere)

# Extract coefficients
slopes     <- c(coef(lm_gla)[2], coef(lm_edi)[2], coef(lm_else)[2])
intercepts <- c(coef(lm_gla)[1], coef(lm_edi)[1], coef(lm_else)[1])
r_sq       <- c(summary(lm_gla)$r.squared, summary(lm_edi)$r.squared, summary(lm_else)$r.squared)

# 95% CIs for slopes
ci_gla  <- confint(lm_gla,  "Income_rate", level = 0.95)
ci_edi  <- confint(lm_edi,  "Income_rate", level = 0.95)
ci_else <- confint(lm_else, "Income_rate", level = 0.95)

reg_table <- data.frame(
  Region    = c("Glasgow", "Edinburgh", "Elsewhere"),
  Intercept = round(intercepts, 4),
  Slope     = round(slopes, 4),
  CI_lower  = round(c(ci_gla[1], ci_edi[1], ci_else[1]), 4),
  CI_upper  = round(c(ci_gla[2], ci_edi[2], ci_else[2]), 4),
  R_squared = round(r_sq, 4)
)

knitr::kable(reg_table,
             caption = "Regional regression models: Attendance $\\sim$ Income\\_rate with 95\\% CI for slope.",
             align = "c",
             col.names = c("Region", "$\\hat{\\beta}_0$", "$\\hat{\\beta}_1$",
                           "CI lower", "CI upper", "$R^2$"))
```

The slope in Glasgow (`r round(slopes[1], 3)`) is shallower than in Edinburgh (`r round(slopes[2], 3)`) and Elsewhere (`r round(slopes[3], 3)`). Comparing the 95% confidence intervals for $\hat{\beta}_1$: if they do not overlap, this provides evidence that the effect of income deprivation on attendance differs across regions.

```{r scatter_regional, echo=FALSE, fig.height=4, fig.width=6, fig.cap='Regional regression lines for Attendance on Income\\_rate'}

plot(dat$Income_rate, dat$Attendance,
     pch = 16, cex = 0.4,
     col = ifelse(dat$Region == "Glasgow", adjustcolor(col_gla),
           ifelse(dat$Region == "Edinburgh", adjustcolor(col_edi),
                  adjustcolor(col_else))),
     xlab = "Income deprivation rate",
     ylab = "Attendance (proportion)",
     main = "Attendance vs Income Rate by Region")
abline(lm_gla,  col = col_gla,  lwd = 2)
abline(lm_edi,  col = col_edi,  lwd = 2)
abline(lm_else, col = col_else, lwd = 2)
legend("topright",
       legend = c("Glasgow", "Edinburgh", "Elsewhere"),
       col = c(col_gla, col_edi, col_else),
       lty = 1, lwd = 2, cex = 0.8, bty = "n")
```

Figure 7 confirms that Glasgow's regression line is flatter: for a given level of income deprivation, Glasgow's predicted attendance does not drop as steeply as in other regions.

**Prediction intervals.** We use each regional model to predict Attendance for a data zone with Income\_rate $= 0.15$ (close to the overall median).

```{r prediction_intervals, echo=FALSE}

new_data <- data.frame(Income_rate = 0.15)

pred_gla  <- predict(lm_gla,  newdata = new_data, interval = "prediction", level = 0.95)
pred_edi  <- predict(lm_edi,  newdata = new_data, interval = "prediction", level = 0.95)
pred_else <- predict(lm_else, newdata = new_data, interval = "prediction", level = 0.95)

pred_table <- data.frame(
  Region = c("Glasgow", "Edinburgh", "Elsewhere"),
  Fitted = round(c(pred_gla[1], pred_edi[1], pred_else[1]), 4),
  Lower  = round(c(pred_gla[2], pred_edi[2], pred_else[2]), 4),
  Upper  = round(c(pred_gla[3], pred_edi[3], pred_else[3]), 4)
)

knitr::kable(pred_table,
             caption = "95\\% prediction intervals for Attendance at Income\\_rate $= 0.15$.",
             align = "c",
             col.names = c("Region", "Fitted", "Lower", "Upper"))
```

At the same level of income deprivation, the predicted Attendance and its uncertainty differ notably across regions. Glasgow's prediction interval is wider, reflecting its greater residual variability — further suggesting that income deprivation alone does not fully explain Glasgow's attendance patterns.

## 3.4 Proportion Analysis

We now define a binary indicator for "low attendance" (Attendance below the overall sample median) and test whether the proportion of low-attendance data zones differs between regions.

```{r proportion_test, echo=FALSE}

overall_median <- median(dat$Attendance)
dat$low_att <- ifelse(dat$Attendance < overall_median, 1, 0)

# Counts per region
x_gla  <- sum(glasgow$Attendance < overall_median)
x_edi  <- sum(edinburgh$Attendance < overall_median)
x_else <- sum(elsewhere$Attendance < overall_median)

p_gla  <- x_gla / n_gla
p_edi  <- x_edi / n_edi
p_else <- x_else / n_else

prop_summary <- data.frame(
  Region     = c("Glasgow", "Edinburgh", "Elsewhere"),
  N          = c(n_gla, n_edi, n_else),
  Low_count  = c(x_gla, x_edi, x_else),
  Proportion = round(c(p_gla, p_edi, p_else), 4)
)

knitr::kable(prop_summary,
             caption = paste0("Proportion of data zones with low attendance (Attendance $<$ ",
                              round(overall_median, 4), ")."),
             align = "c",
             col.names = c("Region", "N", "Low attendance count", "Proportion"))

# Two-sample z-test: Glasgow vs Elsewhere (manual, Ch.16 style)
p_pool <- (x_gla + x_else) / (n_gla + n_else)
z_stat <- (p_gla - p_else) / sqrt(p_pool * (1 - p_pool) * (1/n_gla + 1/n_else))
p_val_z <- 2 * pnorm(abs(z_stat), lower.tail = FALSE)

# Verify with prop.test
prop_result <- prop.test(c(x_gla, x_else), c(n_gla, n_else), correct = FALSE)

z_table <- data.frame(
  Comparison = "Glasgow vs Elsewhere",
  z_stat     = round(z_stat, 4),
  p_value    = formatC(p_val_z, format = "e", digits = 4)
)

knitr::kable(z_table,
             caption = "Two-sample z-test for difference in proportions of low attendance.",
             align = "c",
             col.names = c("Comparison", "z-statistic", "p-value"))
```

`r round(p_gla * 100, 1)`% of Glasgow's data zones have below-median attendance, compared to `r round(p_else * 100, 1)`% in the rest of Scotland. The two-sample $z$-test yields $z =$ `r round(z_stat, 3)` with $p =$ `r formatC(p_val_z, format = "e", digits = 3)`, confirming that Glasgow has a significantly higher proportion of low-attendance data zones.

## 3.5 Chi-Squared Test for Independence

Finally, we classify Attendance into three categories (Low, Medium, High) using tercile boundaries and test whether the distribution of these categories is independent of region.

```{r chisq_test, echo=FALSE}

# Create attendance categories using terciles
q <- quantile(dat$Attendance, probs = c(1/3, 2/3))
dat$Att_cat <- cut(dat$Attendance,
                   breaks = c(-Inf, q[1], q[2], Inf),
                   labels = c("Low", "Medium", "High"))

# Contingency table
cont_table <- table(dat$Region, dat$Att_cat)

knitr::kable(cont_table,
             caption = "Contingency table: Region $\\times$ Attendance category.")

# Chi-squared test for independence
chi_result <- chisq.test(cont_table)
```

Under $H_0$: Attendance category and Region are independent. The $\chi^2$ test statistic is `r round(chi_result$statistic, 3)` with `r chi_result$parameter` degrees of freedom, yielding $p =$ `r formatC(chi_result$p.value, format = "e", digits = 3)`. Since $p < 0.05$, we reject $H_0$: the distribution of attendance levels is significantly associated with region. Glasgow is disproportionately represented in the Low category, while Edinburgh dominates the High category.

## 3.6 Summary of Section 3

The analysis reveals four key findings:

1. **Regional differences persist beyond Glasgow vs Edinburgh.** Attendance differs significantly across all three regions (pairwise $t$-tests, $p < 0.05$), with Glasgow lowest and Edinburgh highest.
2. **The Income–Attendance relationship varies by region.** Glasgow's regression slope is shallower and its $R^2$ (`r round(r_sq[1], 3)`) is lower than Edinburgh's (`r round(r_sq[2], 3)`) or Elsewhere (`r round(r_sq[3], 3)`), suggesting that income deprivation alone explains less of the variation in Glasgow.
3. **Glasgow has a disproportionate share of low-attendance data zones.** A proportion $z$-test confirms this ($p =$ `r formatC(p_val_z, format = "e", digits = 3)`).
4. **Attendance category is not independent of region.** A $\chi^2$ test ($p =$ `r formatC(chi_result$p.value, format = "e", digits = 3)`) confirms a significant association between region and attendance level.

These results suggest that while income deprivation is an important universal predictor of attendance, **its explanatory power varies geographically**. In Glasgow, other unmeasured factors — such as local education policies, community-level interventions, or housing conditions — may play a comparatively larger role. This "Glasgow effect" warrants further investigation with additional variables.
