---
title: "An Analysis of School Attendance Rates across Scottish Data Zones"
date: ''
output:
  pdf_document: default
  bookdown::html_document2:
    number_sections: false
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
dat <- read.csv("../scottishData.csv")

# Unified colour palette
col_gla   <- "#D6604D"   # terracotta  — Glasgow
col_edi   <- "#4393C3"   # steel blue  — Edinburgh
col_else  <- "#66C2A5"   # mint green  — Elsewhere
col_point <- "#525252"   # slate gray  — scatter points
col_line  <- "#2166AC"   # deep blue   — regression line
col_ref   <- "#878787"   # medium gray — reference line
```


# Section 2: Correlation and Linear Regression

Having established that Attendance differs between Glasgow and Edinburgh, we now consider all 400 data zones together to investigate which socio-economic factors are most closely associated with Attendance. We select four candidate predictors: Attainment, crime_rate, CIF (Community Integration Factor), and Income_rate.

## 2.1 Pairwise Relationships

```{r scatters_1, echo=FALSE, fig.height=4, fig.width=6, fig.cap='Pairwise scatterplot matrix of Attendance and four selected variables'}

vars <- c("Attendance", "Attainment", "crime_rate", "CIF", "Income_rate")

pairs(dat[, vars],
      pch = 16, cex = 0.4, col = adjustcolor(col_point, alpha.f = 0.5),
      main = "",
      labels = c("Attendance", "Attainment", "Crime Rate", "CIF", "Income Rate"))
```

Figure 3 provides an overview of the bivariate relationships. Attendance appears negatively associated with Income_rate, CIF, and crime_rate, while displaying a positive trend with Attainment. The strongest visual pattern is the tight negative relationship between Attendance and Income_rate. To quantify these associations, we compute the full matrix of Pearson correlation coefficients.

## 2.2 Pearson Correlation Coefficients

```{r Pairwise Pearson correlation coefficients, echo = FALSE}
# Full 5x5 pairwise Pearson correlation matrix
cor_matrix <- round(cor(dat[, vars], use = "complete.obs", method = "pearson"), 3)
colnames(cor_matrix) <- c("Attendance", "Attainment", "Crime Rate", "CIF", "Income Rate")
rownames(cor_matrix) <- c("Attendance", "Attainment", "Crime Rate", "CIF", "Income Rate")

knitr::kable(cor_matrix,
             caption = "Pairwise Pearson correlation coefficients between the five selected variables.",
             align = "c")

# Store individual correlations for later use
r_attainment <- cor_matrix[1, 2]
r_crime      <- cor_matrix[1, 3]
r_cif        <- cor_matrix[1, 4]
r_income     <- cor_matrix[1, 5]
```

Table 6 confirms the visual impression. Attendance is most strongly correlated with Income_rate ($r =$ `r r_income`), followed by CIF ($r =$ `r r_cif`) and Attainment ($r =$ `r r_attainment`). Crime_rate shows a weaker association ($r =$ `r r_crime`). Notably, Income_rate, CIF, and crime_rate are also highly inter-correlated, reflecting the well-known overlap between deprivation indicators.

Since Income_rate has the highest absolute correlation with Attendance, we select it as the predictor for a simple linear regression model.

## 2.3 Simple Linear Regression

```{r Linear Regression, echo = FALSE}
# Identify the variable with the highest absolute correlation with Attendance
correlations <- c(
  Attainment = r_attainment,
  Crime_rate = r_crime,
  CIF = r_cif,
  Income_rate = r_income
)
data2_name <- names(which.max(abs(correlations)))

# Manual calculation of slope and intercept (Workshop 5 style)
x <- dat[[data2_name]]
y <- dat$Attendance
n <- length(x)
beta1_manual <- (sum(x * y) - n * mean(x) * mean(y)) / (sum(x^2) - n * mean(x)^2)
beta0_manual <- mean(y) - beta1_manual * mean(x)

# Verify with lm()
model_formula <- as.formula(paste("Attendance ~", data2_name))
lm_model <- lm(model_formula, data = dat)

# Extract coefficients
intercept <- coef(lm_model)[1]
slope     <- coef(lm_model)[2]
r_squared <- summary(lm_model)$r.squared
```

The variable with the highest absolute correlation with Attendance is **`r data2_name`** ($r =$ `r round(correlations[data2_name], 3)`). We therefore fit a simple linear regression model:
$$\text{Attendance} = \beta_0 + \beta_1 \times \text{`r gsub("_", "\\\\_", data2_name)`} + \varepsilon$$

The estimated regression equation is: $\hat{y} =$ `r round(intercept, 4)` $+$ (`r round(slope, 4)`) $\times x$.

```{r Linear Regression Analysis, echo = FALSE}

# Regression coefficients table
coef_table <- summary(lm_model)$coefficients
knitr::kable(coef_table, digits = 4,
             caption = paste("Simple linear regression model: Attendance ~", data2_name))

# 95% confidence intervals for coefficients
ci <- confint(lm_model, level = 0.95)
rownames(ci) <- c("Intercept ($\\beta_0$)", "Slope ($\\beta_1$)")
colnames(ci) <- c("2.5%", "97.5%")
knitr::kable(ci, digits = 4, caption = "95% Confidence Intervals for Regression Coefficients")
```

Both coefficients are highly significant ($p < 0.001$). The slope $\hat{\beta}_1 =$ `r round(slope, 4)` indicates that, on average, a one-unit increase in `r data2_name` is associated with a `r round(abs(slope), 4)` decrease in Attendance. The model explains `r round(r_squared * 100, 2)`% of the total variance ($R^2 =$ `r round(r_squared, 4)`), suggesting that `r data2_name` alone is a strong predictor of Attendance.

## 2.4 Visualisation of the Fitted Model

```{r linear_regression, echo = FALSE, fig.height=4, fig.width=6, fig.cap='Scatterplot with fitted regression line'}

equation <- paste0("y = ", round(intercept, 3), " + (", round(slope, 3), ") x")
r2_label <- paste0("R² = ", round(r_squared, 4))

plot(dat[[data2_name]], dat$Attendance,
     pch = 16, cex = 0.5, col = adjustcolor(col_point, 0.5),
     xlab = data2_name, ylab = "Attendance (proportion)",
     main = paste("Attendance vs", data2_name))
abline(lm_model, col = col_line, lwd = 2)
legend("topright", legend = c(equation, r2_label),
       col = c(col_line, NA), lty = c(1, NA), lwd = c(2, NA),
       cex = 0.85, bty = "n")

```

Figure 4 visualises the fitted regression line superimposed on the raw data. The negative trend is clear. Some dispersion is evident at higher values of `r data2_name`, which we examine further via residual diagnostics.

## 2.5 Residual Diagnostics

```{r residuals, echo = FALSE, fig.height=4, fig.width=6, fig.cap='Residuals of the fitted model'}

# residuals (Workshop 5 style)
dat$residuals <- residuals(lm_model)
dat$fitted    <- fitted(lm_model)

plot(dat$fitted, dat$residuals,
     pch = 16, cex = 0.5, col = adjustcolor(col_point, 0.5),
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted Values")
abline(h = 0, lty = 2, col = col_ref, lwd = 1.5)

```

The residual plot (Figure 5) shows residuals scattered around zero with no obvious systematic pattern or funnel shape, indicating that the linearity and constant-variance assumptions are reasonably satisfied. There are a few large negative residuals (data zones with unusually low attendance given their income deprivation level), which could be investigated further.

Overall, the results of Section 2 demonstrate that income deprivation is the single strongest linear predictor of attendance among the variables considered, explaining approximately `r round(r_squared * 100, 0)`% of the variation. This complements the city-level differences found in Section 1 and suggests that socio-economic deprivation may be a key driver of the lower attendance rates observed in Glasgow. An important follow-up question is whether this relationship holds equally across regions — we investigate this in Section 3.